---
title: "Wetland_local_models"
author: "Michael Krochta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

*General setup*
```{r}
library(tidyverse)
library(lubridate)
library(caret)
library(ranger)
library(xgboost)
library(doParallel)

# Paths 
data_dir    <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/local_mod_data"
results_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results"
```

*Data setup*
```{r}
# Helpers 
# Robust date parser
fix_dates <- function(df) {
  if (!inherits(df$Datetime, "Date")) {
    df$Datetime <- as.Date(
      df$Datetime,
      tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y",
                     "%Y-%m-%d %H:%M:%S","%m/%d/%Y %H:%M","%m/%d/%Y %H:%M:%S")
    )
  }
  df
}

# Parse site + metric from filename like "Anvil_max_WT_model_data.csv"
parse_site_metric <- function(path) {
  f <- basename(path)
  m <- regexec("^([A-Za-z0-9]+)_(max|mean|range)_WT_model_data\\.csv$", f)
  parts <- regmatches(f, m)[[1]]
  if (length(parts) != 3) stop("Unexpected filename pattern: ", f)
  list(site = parts[2], metric_key = parts[3])  # metric_key in {max, mean, range}
}

# Response column map
resp_cols <- list(max = "max_WT", mean = "mean_WT", range = "range_WT")

# Remote-only exclusion (match your global setup)
remote_exclude_exact <- c(
  "WL_daily_mean","WL_daily_max","WL_daily_min","WL_daily_range",
  "WL_mean_2day","WL_mean_3day","WL_mean_7day",
  "WL_max_2day","WL_max_3day","WL_max_7day",
  "WL_min_2day","WL_min_3day","WL_min_7day",
  "GWT_mean","GWT_max","GWT_min","GWT_range",
  "GWT_mean_2day","GWT_mean_3day","GWT_mean_7day",
  "GWT_max_2day","GWT_max_3day","GWT_max_7day",
  "GWT_min_2day","GWT_min_3day","GWT_min_7day",
  "GWL_mean","GWL_max","GWL_min","GWL_range",
  "GWL_mean_2day","GWL_mean_3day","GWL_mean_7day",
  "GWL_max_2day","GWL_max_3day","GWL_max_7day",
  "GWL_min_2day","GWL_min_3day","GWL_min_7day",
  "Water_area_m2_3d"
)
remote_exclude_prefix <- c("WL_", "GWT_", "GWL_", "Water")  # keep as in your global pipeline

get_predictors <- function(df, response_col, mode = c("all","remote")) {
  mode <- match.arg(mode)
  base <- setdiff(names(df), c(response_col, "Datetime", "Site"))  # never model on ID/time
  if (mode == "all") return(base)
  drop_prefixed <- unlist(lapply(remote_exclude_prefix, function(p) grep(paste0("^", p), base, value = TRUE)))
  setdiff(base, unique(c(remote_exclude_exact, drop_prefixed)))
}

```


*5-fold CV fxn*
```{r}
# Weekly-blocked CV with automatic k based on available weeks
make_weekly_folds <- function(df, k = 5, date_col = "Datetime") {
  df <- df %>% mutate(.week = lubridate::floor_date(.data[[date_col]], unit = "week", week_start = 1))
  weeks <- sort(unique(df$.week))
  k_eff <- min(k, length(weeks))
  if (k_eff < 2) stop(sprintf("Not enough unique weeks (%d) for CV (need >= 2).", length(weeks)))
  set.seed(42)
  weeks <- sample(weeks)
  folds <- split(weeks, cut(seq_along(weeks), breaks = k_eff, labels = FALSE))
  index <- list(); indexOut <- list()
  for (i in seq_along(folds)) {
    val_weeks <- folds[[i]]
    val_idx <- which(df$.week %in% val_weeks)
    trn_idx <- setdiff(seq_len(nrow(df)), val_idx)
    index[[paste0("Fold", i)]]    <- trn_idx
    indexOut[[paste0("Fold", i)]] <- val_idx
  }
  list(index = index, indexOut = indexOut, k = length(folds))
}
```

*Performance metrics*
```{r}
# Metrics 
RMSE_ <- function(pred, obs) sqrt(mean((pred - obs)^2, na.rm = TRUE))
MAE_  <- function(pred, obs) mean(abs(pred - obs), na.rm = TRUE)
nse   <- function(obs, pred) 1 - sum((obs - pred)^2, na.rm = TRUE) /
                              sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)

compute_metrics <- function(pred_df, truth_col = "obs", pred_col = "pred", site_col = "Site") {
  overall <- tibble(
    RMSE = RMSE_(pred_df[[pred_col]], pred_df[[truth_col]]),
    MAE  = MAE_ (pred_df[[pred_col]], pred_df[[truth_col]]),
    R2   = caret::R2(pred_df[[pred_col]], pred_df[[truth_col]]),
    NSE  = nse  (pred_df[[truth_col]], pred_df[[pred_col]])
  )
  by_site <- pred_df %>%
    group_by(.data[[site_col]]) %>%
    summarise(
      RMSE = RMSE_(.data[[pred_col]], .data[[truth_col]]),
      MAE  = MAE_ (.data[[pred_col]], .data[[truth_col]]),
      R2   = caret::R2(.data[[pred_col]], .data[[truth_col]]),
      NSE  = nse  (.data[[truth_col]], .data[[pred_col]]),
      .groups = "drop"
    ) %>% rename(Site = .data[[site_col]])
  list(overall = overall, by_site = by_site)
}
```

*Hyperparameters*
```{r}
# Hyperparameter grids 
rf_grid_refined <- function(p) {
  expand.grid(
    mtry          = sort(unique(pmax(1, c(
      floor(sqrt(p))-1, floor(sqrt(p)), floor(sqrt(p))+1,
      floor(p/4), floor(p/3), floor(p/2), ceiling(2*p/3)
    )))),
    splitrule     = "variance",
    min.node.size = c(1, 2, 3, 5, 7, 10, 15),
    stringsAsFactors = FALSE
  )
}

xgb_grid_refined <- expand.grid(
  nrounds          = c(400, 800),
  eta              = c(0.05, 0.1),
  max_depth        = c(6, 9, 12),
  min_child_weight = c(1, 4),
  subsample        = c(0.8, 0.9),
  colsample_bytree = c(0.8, 1.0),
  gamma            = c(0, 0.1, 1.0)
)
```


*Train models, prepping data*
```{r}
# Core trainer (unchanged logic from global; now used on single-site data)
train_model <- function(data,
                        response,
                        predictors,
                        model_type = c("xgb", "rfr"),
                        k = 5,
                        site_col = "Site",
                        date_col = "Datetime",
                        tune_grid = NULL) {

  model_type <- match.arg(model_type)

  # Keep only needed columns; drop rows with NA
  df <- data %>% dplyr::select(all_of(c(site_col, date_col, response, predictors))) %>% na.omit()

  # Build blocked folds (weekly) with auto k
  folds <- make_weekly_folds(df, k = k, date_col = date_col)

  # caret control
  ctrl <- trainControl(
    method          = "cv",
    number          = folds$k,
    index           = folds$index,
    indexOut        = folds$indexOut,
    verboseIter     = FALSE,
    savePredictions = "final",
    allowParallel   = TRUE
  )

  # Tuning grid
  if (is.null(tune_grid)) {
    tune_grid <- if (model_type == "rfr") rf_grid_refined(length(predictors)) else xgb_grid_refined
  }

  # Train
  form <- as.formula(paste(response, "~ ."))
  train_df <- df %>% dplyr::select(all_of(c(response, predictors)))

  set.seed(2025)
  if (model_type == "rfr") {
    fit <- train(
      form, data = train_df, method = "ranger",
      trControl = ctrl, tuneGrid = tune_grid, metric = "RMSE",
      importance = "permutation",
      num.trees = 500,
      num.threads = 1     # avoid nested parallel
    )
  } else {
    fit <- train(
      form, data = train_df, method = "xgbTree",
      trControl = ctrl, tuneGrid = tune_grid, metric = "RMSE",
      nthread = 1         # avoid nested parallel
    )
  }

  # Align OOF predictions back to Site/Date
  if (!"rowIndex" %in% names(fit$pred)) stop("caret did not return 'rowIndex' in fit$pred.")
  oof <- fit$pred %>%
    mutate(
      Site = df[[site_col]][rowIndex],
      Date = df[[date_col]][rowIndex]
    ) %>%
    transmute(Site, Date, obs = obs, pred = pred)

  # Metrics
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # Var importance
  vimp <- varImp(fit)$importance %>%
    tibble::rownames_to_column("Variable") %>%
    arrange(desc(Overall))

  list(
    response        = response,
    model_type      = model_type,
    predictors      = predictors,
    best_params     = fit$bestTune,
    cv_results      = fit$results,
    metrics_overall = mets$overall,
    metrics_by_site = mets$by_site,
    var_importance  = vimp,
    oof_predictions = oof,
    fit_object      = fit
  )
}

# Build datasets for a single local file (Set A = all, Set B = remote-only)
prep_dataset_local <- function(df, response_key, site_name, mode = c("all","remote")) {
  mode <- match.arg(mode)
  response_col <- resp_cols[[response_key]]
  if (!response_col %in% names(df)) stop("Response column not found: ", response_col)
  df2 <- df %>% mutate(Site = site_name)
  preds <- get_predictors(df2, response_col, mode = mode)
  list(data = df2, response = response_col, predictors = preds)
}

```

*Model run fxn*
```{r}
# Run all local files 
run_all_local <- function(data_dir, k = 5) {
  files <- list.files(data_dir, pattern = "_(max|mean|range)_WT_model_data\\.csv$", full.names = TRUE)
  if (length(files) == 0) stop("No local files found in: ", data_dir)

  results <- list()

  # parallel outer loop
  cl <- makeCluster(parallel::detectCores() - 1)
  registerDoParallel(cl)

  on.exit({
    stopCluster(cl)
    registerDoSEQ()
  }, add = TRUE)

  for (fp in files) {
    info <- parse_site_metric(fp)
    site <- info$site
    resp_key <- info$metric_key
    response_col <- resp_cols[[resp_key]]

    message(sprintf(">>> %s | %s_WT", site, resp_key))

    df <- read.csv(fp, stringsAsFactors = FALSE) %>% fix_dates()

    # --- Set A: All predictors (single-site file; no site exclusion)
    A <- prep_dataset_local(df, resp_key, site, mode = "all")
    # XGB
    results[[paste(site, resp_key, "A_xgb", sep = "_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "xgb", k = k
    )
    # RFR
    results[[paste(site, resp_key, "A_rfr", sep = "_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "rfr", k = k
    )

    # --- Set B: Remote-only
    B <- prep_dataset_local(df, resp_key, site, mode = "remote")
    if (length(B$predictors) >= 1) {  # skip if remote-only leaves nothing
      results[[paste(site, resp_key, "B_xgb", sep = "_")]] <- train_model(
        data = B$data, response = B$response, predictors = B$predictors,
        model_type = "xgb", k = k
      )
      results[[paste(site, resp_key, "B_rfr", sep = "_")]] <- train_model(
        data = B$data, response = B$response, predictors = B$predictors,
        model_type = "rfr", k = k
      )
    } else {
      warning(sprintf("No predictors left for remote-only at %s %s; skipping Set B.", site, resp_key))
    }
  }

  results
}

```

*Run models, get outputs*
```{r}
# Execute 
local_results_refined <- run_all_local(data_dir, k = 5)

# Save full object
saveRDS(local_results_refined, file.path(results_dir, "local_models_refined.rds"))

# Summarize overall metrics
summary_tbl <- purrr::imap_dfr(local_results_refined, function(res, key) {
  parts <- strsplit(key, "_")[[1]]
  tibble(
    Site   = parts[1],
    Metric = parts[2],           # max / mean / range
    Set    = parts[3],           # A / B
    Model  = toupper(parts[4]),  # xgb / rfr
    RMSE   = res$metrics_overall$RMSE,
    MAE    = res$metrics_overall$MAE,
    R2     = res$metrics_overall$R2,
    NSE    = res$metrics_overall$NSE,
    n_pred = length(res$predictors)
  )
}) %>% arrange(Site, Metric, Set, Model)

readr::write_csv(summary_tbl, file.path(results_dir, "local_models_refined_summary.csv"))

message("Done. Results:\n  - RDS: ", file.path(results_dir, "local_models_refined.rds"),
        "\n  - CSV: ", file.path(results_dir, "local_models_refined_summary.csv"))
```

*Evaluating results*
```{r}

# paths
results_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results"
summary_path <- file.path(results_dir, "local_models_refined_summary.csv")

# 1) load your existing summary (overall OOF metrics per run)
summary_tbl <- readr::read_csv(summary_path, show_col_types = FALSE)
# Expecting columns like: Site, Metric, Set, Model, RMSE, MAE, R2, NSE, n_pred

# helper to parse key -> fields
parse_key <- function(key){
  parts <- strsplit(key, "_")[[1]]
  tibble(Site = parts[1], Metric = parts[2], Set = parts[3], Model = toupper(parts[4]))
}

# 2) build small add-on tables from local_results_refined
# n_oof per run
n_oof_tbl <- purrr::imap_dfr(local_results_refined, function(res, key){
  parse_key(key) %>% mutate(n_oof = nrow(res$oof_predictions))
})

# resample diagnostics
resample_diag <- purrr::imap_dfr(local_results_refined, function(res, key){
  rs <- res$fit_object$resample
  parse_key(key) %>%
    mutate(
      n_resamples = nrow(rs),
      rmse_mean   = mean(rs$RMSE, na.rm = TRUE),
      rmse_sd     = sd(rs$RMSE,   na.rm = TRUE),
      na_RMSE     = sum(is.na(rs$RMSE)),
      na_R2       = sum(is.na(rs$Rsquared))
    )
})

# best-tuned hyperparameters (wide)
best_params_wide <- purrr::imap_dfr(local_results_refined, function(res, key){
  bp <- as_tibble(res$best_params); if (ncol(bp) == 0) bp <- tibble()
  bind_cols(parse_key(key), bp)
})

# edge-of-grid flag
edge_flag <- purrr::imap_dfr(local_results_refined, function(res, key){
  bt <- res$best_params; grid <- res$cv_results
  at_edge_any <- FALSE
  if (!is.null(bt) && !is.null(grid) && nrow(grid) > 0) {
    ps <- intersect(names(bt), names(grid))
    at_edge_any <- any(purrr::map_lgl(ps, function(p){
      col <- grid[[p]]; val <- bt[[p]]
      if (is.factor(col)) col <- as.character(col)
      if (is.numeric(col)) {
        suppressWarnings(as.numeric(val)) %in% range(col, na.rm = TRUE)
      } else {
        u <- sort(unique(as.character(col)))
        length(u) > 1 && as.character(val) %in% c(u[1], u[length(u)])
      }
    }))
  }
  parse_key(key) %>% mutate(edge_hit = at_edge_any)
})

# 3) join add-ons onto your existing summary (no duplication)
local_models_master <- summary_tbl %>%
  left_join(n_oof_tbl,      by = c("Site","Metric","Set","Model")) %>%
  left_join(resample_diag,  by = c("Site","Metric","Set","Model")) %>%
  left_join(best_params_wide,by = c("Site","Metric","Set","Model")) %>%
  left_join(edge_flag,      by = c("Site","Metric","Set","Model")) %>%
  mutate(retune = (edge_hit %in% TRUE) | (na_RMSE > 0)) %>%
  arrange(Site, Metric, Set, Model)

# 4) write the consolidated table
out_path <- file.path(results_dir, "local_models_master.csv")
readr::write_csv(local_models_master, out_path)

message("Wrote: ", out_path)

```

*Evaluating hyperparameters and k-folds and re-running individual models where needed*
```{r}

data_dir    <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/local_mod_data"
results_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results_pass2"
master_path <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_master.csv"
compare_out <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_master_pass2_compare.csv"

dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

# Load master & normalize / ensure columns 
master <- readr::read_csv(master_path, show_col_types = FALSE) %>%
  mutate(Model = toupper(Model))

if (!"edge_hit" %in% names(master)) master$edge_hit <- FALSE
if (!"na_RMSE"  %in% names(master)) master$na_RMSE  <- 0

if (!"cv_instability" %in% names(master)) {
  if (all(c("rmse_sd","rmse_mean") %in% names(master))) {
    master <- master %>%
      mutate(cv_instability = ifelse(is.finite(rmse_mean) & rmse_mean > 0,
                                     (rmse_sd / rmse_mean) > 0.35, FALSE))
  } else {
    master$cv_instability <- FALSE
  }
}

master <- master %>%
  mutate(
    edge_hit       = replace_na(edge_hit, FALSE),
    cv_instability = replace_na(cv_instability, FALSE),
    na_RMSE        = replace_na(na_RMSE, 0)
  )

# Edge-at-top checks (what *really* warrants expansion) 
edge_high_xgb <- function(row) {
  isTRUE(row$nrounds == 800) || isTRUE(row$max_depth == 12)
}
edge_high_rfr <- function(row) {
  mns <- dplyr::coalesce(row$`min.node.size`, row$min.node.size)
  high_mns  <- isTRUE(mns == 15)
  high_mtry <- !is.na(row$mtry) && !is.na(row$n_pred) && row$mtry >= ceiling(0.75 * row$n_pred)
  high_mns || high_mtry
}

master <- master %>%
  rowwise() %>%
  mutate(edge_high_core = if (Model == "XGB") edge_high_xgb(cur_data()) else edge_high_rfr(cur_data())) %>%
  ungroup()

# Choose pass-2 targets (keep ALL runs; only refit flagged ones) 
targets <- master %>%
  filter(NSE < 0 | cv_instability | na_RMSE > 0 | edge_high_core) %>%
  arrange(Site, Metric, Set, Model)

# Preview (optional)
targets %>%
  select(Site, Metric, Set, Model, RMSE, NSE, edge_high_core, cv_instability, na_RMSE)
message("Targets to refit: ", nrow(targets))

# Micro-grids: tweak only edges; keep other params fixed 
expand_grid_xgb_min <- function(row){
  nr <- ifelse(is.na(row$nrounds), 800L, as.integer(row$nrounds))
  md <- ifelse(is.na(row$max_depth), 9L,  as.integer(row$max_depth))
  mc <- ifelse(is.na(row$min_child_weight), 1L, as.integer(row$min_child_weight))
  et <- ifelse(is.na(row$eta), 0.05, as.numeric(row$eta))
  ss <- ifelse(is.na(row$subsample), 0.8, as.numeric(row$subsample))
  cs <- ifelse(is.na(row$colsample_bytree), 0.8, as.numeric(row$colsample_bytree))
  ga <- ifelse(is.na(row$gamma), 0.0, as.numeric(row$gamma))

  edge  <- isTRUE(row$edge_high_core)
  shaky <- isTRUE(row$cv_instability) || (is.numeric(row$na_RMSE) && row$na_RMSE > 0) || (is.numeric(row$NSE) && row$NSE < 0)

  nr_vals <- if (edge && nr >= 800)            unique(c(nr, nr+200, nr+400))
        else if (shaky)                         unique(c(max(200L, nr-200L), nr, nr+200L))
        else                                    nr

  md_vals <- if (edge && md >= 12)             unique(c(md, md+2))
        else if (shaky)                         unique(c(max(2L, md-2L), md, md+2L))
        else                                    md

  mc_vals <- if (mc %in% c(1,4))               unique(c(mc, max(1, mc-1), mc+1)) else mc
  ga_vals <- if (ga %in% c(0,1.0))             unique(c(ga, ifelse(ga==0, 0.05, 2.0))) else ga

  expand.grid(
    nrounds          = nr_vals,
    max_depth        = md_vals,
    min_child_weight = mc_vals,
    gamma            = ga_vals,
    eta              = et,
    subsample        = ss,
    colsample_bytree = cs,
    KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
  )
}

expand_grid_rfr_min <- function(row){
  p  <- ifelse(is.na(row$n_pred), NA_integer_, as.integer(row$n_pred))
  mt <- ifelse(is.na(row$mtry), ifelse(is.na(p), 3L, max(1L, floor(sqrt(p)))), as.integer(row$mtry))
  mns <- dplyr::coalesce(row$`min.node.size`, row$min.node.size)
  mns <- ifelse(is.na(mns), 5L, as.integer(mns))

  edge  <- isTRUE(row$edge_high_core)
  shaky <- isTRUE(row$cv_instability) || (is.numeric(row$na_RMSE) && row$na_RMSE > 0) || (is.numeric(row$NSE) && row$NSE < 0)

  mtry_vals <-
    if (!is.na(p) && edge && mt >= ceiling(0.75*p)) sort(unique(pmin(p, c(mt, ceiling(0.85*p), ceiling(0.95*p))))) else
    if (!is.na(p) && shaky)                         sort(unique(c(max(1L, mt-1L), mt, min(p, mt+1L)))) else
                                                     mt

  mns_vals <- if (mns %in% c(1,15)) unique(c(mns, ifelse(mns==1, 2L, 20L))) else mns

  expand.grid(
    mtry          = mtry_vals,
    splitrule     = "variance",
    min.node.size = mns_vals,
    KEEP.OUT.ATTRS = FALSE, stringsAsFactors = FALSE
  )
}

# Runner: fits ONLY the flagged runs; supports dry-run & resume 
run_pass2_targets <- function(
  targets,
  data_dir    = data_dir,
  results_dir = results_dir,
  execute     = FALSE,
  resume      = TRUE
){
  dir.create(results_dir, recursive = TRUE, showWarnings = FALSE)

  for (i in seq_len(nrow(targets))) {
    row   <- targets[i,]
    site  <- row$Site
    metric<- row$Metric         # "max"/"mean"/"range"
    setAB <- row$Set            # "A"/"B"
    model <- tolower(row$Model) # "xgb"/"rfr"

    in_fp   <- file.path(data_dir, sprintf("%s_%s_WT_model_data.csv", site, metric))
    out_rds <- file.path(results_dir, sprintf("%s_%s_%s_%s_pass2.rds", site, metric, setAB, model))

    if (resume && file.exists(out_rds)) { message("SKIP (exists): ", basename(out_rds)); next }

    tg <- if (model == "xgb") expand_grid_xgb_min(row) else expand_grid_rfr_min(row)
    # Use k=4 only if we previously saw fold issues
    k2 <- if (isTRUE(row$cv_instability) || (is.numeric(row$na_RMSE) && row$na_RMSE > 0)) 4L else 5L

    message(sprintf("%s | %s_WT | Set %s | %s | k=%d | grid=%d",
                    site, metric, setAB, toupper(model), k2, nrow(tg)))

    if (!execute) next  # DRY RUN only

    # Load + prep (uses your existing helpers)
    df <- read.csv(in_fp, stringsAsFactors = FALSE)
    if (!inherits(df$Datetime, "Date")) {
      df$Datetime <- as.Date(df$Datetime,
        tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y",
                       "%Y-%m-%d %H:%M:%S","%m/%d/%Y %H:%M","%m/%d/%Y %H:%M:%S"))
    }
    D <- prep_dataset_local(df, metric, site, mode = ifelse(setAB=="A","all","remote"))
    if (!length(D$predictors)) { warning("No predictors after filtering: ", site, " ", metric, " Set ", setAB); next }

    res2 <- train_model(
      data = D$data, response = D$response, predictors = D$predictors,
      model_type = model, k = k2, tune_grid = tg
    )
    saveRDS(res2, out_rds)
  }
  invisible(NULL)
}

# 
# 1) Dry-run (prints what would be rerun; does NOT refit)
# passing paths explicitly to avoid the default-arg quirk
run_pass2_targets(
  targets,
  data_dir    = "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/local_mod_data",
  results_dir = "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results_pass2",
  execute     = TRUE,   # <-- actually fit now
  resume      = TRUE    # <-- safe to rerun; skips any finished jobs
)


# 2) actually run just those flagged ones
# run_pass2_targets(targets, execute = TRUE)


# PASS-2 COMPARISON (SAFE)

collect_pass2 <- function(dir) {
  files <- list.files(dir, pattern = "_pass2\\.rds$", full.names = TRUE)
  if (!length(files)) return(tibble())  # safe empty
  purrr::map_dfr(files, function(fp){
    fn <- basename(fp)
    m <- regexec("^([A-Za-z0-9]+)_(max|mean|range)_(A|B)_(xgb|rfr)_pass2\\.rds$", fn)
    parts <- regmatches(fn, m)[[1]]
    if (length(parts) != 5) return(NULL)
    Site   <- parts[2]; Metric <- parts[3]; Set <- parts[4]; Model <- toupper(parts[5])
    res <- readRDS(fp)
    tibble(
      Site, Metric, Set, Model,
      RMSE_p2 = res$metrics_overall$RMSE,
      MAE_p2  = res$metrics_overall$MAE,
      R2_p2   = res$metrics_overall$R2,
      NSE_p2  = res$metrics_overall$NSE,
      n_resamples_p2 = nrow(res$fit_object$resample),
      grid_rows_p2   = nrow(res$cv_results),
      pass2_file     = fp
    )
  })
}

pass2 <- collect_pass2(results_dir)

if (nrow(pass2) == 0) {
  message("No pass-2 files found in: ", results_dir,
          "\nRun `run_pass2_targets(targets, execute = TRUE)` first, then re-run this compare.")
} else {
  base <- master %>% 
    select(Site, Metric, Set, Model, RMSE, MAE, R2, NSE, n_resamples)

  compare <- pass2 %>%
    left_join(base, by = c("Site","Metric","Set","Model")) %>%
    mutate(
      dRMSE = RMSE_p2 - RMSE,   # negative = improvement
      dMAE  = MAE_p2  - MAE,
      dR2   = R2_p2   - R2,
      dNSE  = NSE_p2  - NSE
    ) %>%
    arrange(Site, Metric, Set, Model)

  readr::write_csv(compare, compare_out)
  message("Wrote comparison CSV: ", compare_out)
}

```


*Final local models results table*
```{r}
MASTER_CSV <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_master.csv"
PASS2_DIR  <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results_pass2"
FINAL_OUT  <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_final.csv"

#  Load original master & normalize 
orig <- readr::read_csv(MASTER_CSV, show_col_types = FALSE) %>%
  mutate(Model = toupper(Model))

# normalize possible naming differences for RFR:
if ("min_node_size" %in% names(orig) && !"min.node.size" %in% names(orig)) {
  orig <- orig %>% rename(`min.node.size` = min_node_size)
}

# keep only the columns we care about + set k (originals used k=5)
keep_cols <- c("Site","Metric","Set","Model","RMSE","MAE","R2","NSE",
               # XGB params (may be NA for RFR rows)
               "nrounds","eta","max_depth","min_child_weight","subsample","colsample_bytree","gamma",
               # RFR params (may be NA for XGB rows)
               "mtry","min.node.size")
present_cols <- intersect(keep_cols, names(orig))

orig_slim <- orig %>%
  select(any_of(present_cols)) %>%
  mutate(k = 5L, source = "orig")

# Collect pass-2 results from RDS 
collect_pass2 <- function(dir) {
  files <- list.files(dir, pattern = "_pass2\\.rds$", full.names = TRUE)
  if (!length(files)) return(tibble())  # safe empty

  purrr::map_dfr(files, function(fp){
    fn <- basename(fp)
    m <- regexec("^([A-Za-z0-9]+)_(max|mean|range)_(A|B)_(xgb|rfr)_pass2\\.rds$", fn)
    parts <- regmatches(fn, m)[[1]]
    if (length(parts) != 5) return(NULL)

    Site   <- parts[2]
    Metric <- parts[3]
    Set    <- parts[4]
    Model  <- toupper(parts[5])

    res <- readRDS(fp)

    # metrics
    mets <- res$metrics_overall %>%
      transmute(RMSE = RMSE, MAE = MAE, R2 = R2, NSE = NSE)

    # best params (caret bestTune)
    bp <- tibble::as_tibble(res$best_params)

    # ensure all expected param cols exist (fill with NA if missing)
    xgb_cols <- c("nrounds","eta","max_depth","min_child_weight","subsample","colsample_bytree","gamma")
    rfr_cols <- c("mtry","min.node.size")
    for (cc in c(xgb_cols, rfr_cols)) {
      if (!cc %in% names(bp)) bp[[cc]] <- NA
    }

    # k via Resample column (Fold1, Fold2, ...)
    k_used <- tryCatch({
      if (!is.null(res$fit_object$resample$Resample)) {
        length(unique(res$fit_object$resample$Resample))
      } else NA_integer_
    }, error = function(e) NA_integer_)

    tibble(
      Site   = Site,
      Metric = Metric,
      Set    = Set,
      Model  = Model,
      RMSE   = mets$RMSE[1],
      MAE    = mets$MAE[1],
      R2     = mets$R2[1],
      NSE    = mets$NSE[1],
      # params:
      nrounds = bp$nrounds[1],
      eta = bp$eta[1],
      max_depth = bp$max_depth[1],
      min_child_weight = bp$min_child_weight[1],
      subsample = bp$subsample[1],
      colsample_bytree = bp$colsample_bytree[1],
      gamma = bp$gamma[1],
      mtry = bp$mtry[1],
      `min.node.size` = bp$`min.node.size`[1],
      k = k_used,
      source = "pass2"
    )
  }) %>% arrange(Site, Metric, Set, Model)
}

p2 <- collect_pass2(PASS2_DIR)

# Compose FINAL table: pass-2 overrides orig where present 
if (nrow(p2) == 0) {
  message("No pass-2 files found; final table = originals only.")
  final <- orig_slim
} else {
  keys <- c("Site","Metric","Set","Model")

  # originals that DO NOT have a pass-2 counterpart
  orig_only <- orig_slim %>%
    anti_join(p2 %>% select(all_of(keys)), by = keys)

  # union with pass-2 rows
  # (both have the same column set except possibly missing some param cols)
  all_cols <- union(names(orig_only), names(p2))
  orig_only <- orig_only %>% mutate(across(setdiff(all_cols, names(.)), ~NA))
  p2        <- p2        %>% mutate(across(setdiff(all_cols, names(.)), ~NA))

  final <- bind_rows(orig_only[, all_cols], p2[, all_cols]) %>%
    arrange(Site, Metric, Set, Model)
}

# Save final table 
readr::write_csv(final, FINAL_OUT)
message("Wrote final models table: ", FINAL_OUT)

# Optional: quick peek at how many came from pass-2
final %>% count(source)

```


**Local GLS Models**
*Data setup*
```{r}
library(nlme)
library(car)         
library(dplyr)
library(tidyr)
library(purrr)
library(lubridate)
library(tibble)

# Remote-only exclusion 
remote_exclude_exact <- c(
  "WL_daily_mean","WL_daily_max","WL_daily_min","WL_daily_range",
  "WL_mean_2day","WL_mean_3day","WL_mean_7day",
  "WL_max_2day","WL_max_3day","WL_max_7day",
  "WL_min_2day","WL_min_3day","WL_min_7day",
  "GWT_mean","GWT_max","GWT_min","GWT_range",
  "GWT_mean_2day","GWT_mean_3day","GWT_mean_7day",
  "GWT_max_2day","GWT_max_3day","GWT_max_7day",
  "GWT_min_2day","GWT_min_3day","GWT_min_7day",
  "GWL_mean","GWL_max","GWL_min","GWL_range",
  "GWL_mean_2day","GWL_mean_3day","GWL_mean_7day",
  "GWL_max_2day","GWL_max_3day","GWL_max_7day",
  "GWL_min_2day","GWL_min_3day","GWL_min_7day", "Water_area_m2_3d"
)
remote_exclude_prefix <- c("WL_", "GWT_", "GWL_", "Water")

# Response column resolver (local files)
resolve_response_col <- function(df, metric = c("max","mean","range")) {
  metric <- match.arg(metric)
  guess <- switch(metric,
                  max   = "max_WT",
                  mean  = "mean_WT",
                  range = "range_WT")
  if (guess %in% names(df)) return(guess)
  # fallback: first column name containing pattern
  pat <- paste0("^", metric, "(_)?WT")
  hit <- grep(pat, names(df), value = TRUE, ignore.case = TRUE)
  if (length(hit)) return(hit[1])
  stop("Could not find response column for metric='", metric, "'.")
}

# Predictors for a given df + mode (all vs remote-only) 
get_predictors_local <- function(df, response_col, mode = c("all","remote")) {
  mode <- match.arg(mode)
  base <- setdiff(names(df), c(response_col, "Site", "Datetime"))
  if (mode == "all") return(base)
  # remote-only: drop exacts and prefixed
  drop_prefixed <- unlist(lapply(remote_exclude_prefix, function(p) grep(paste0("^", p), base, value = TRUE)))
  setdiff(base, unique(c(remote_exclude_exact, drop_prefixed)))
}

# Date fixer 
fix_dates <- function(df, date_col = "Datetime") {
  if (!inherits(df[[date_col]], "Date")) {
    df[[date_col]] <- as.Date(df[[date_col]],
                              tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y",
                                             "%Y-%m-%d %H:%M:%S","%m/%d/%Y %H:%M","%m/%d/%Y %H:%M:%S"))
  }
  df
}
```

*VIF functions*
```{r}
# VIF pruning
vif_prune <- function(train_df, response, preds, thresh = 5) {
  keep <- preds
  dropped <- character(0)

  # Drop constant columns up front
  nzv <- keep[sapply(keep, function(v) sd(train_df[[v]], na.rm = TRUE) == 0)]
  if (length(nzv)) {
    keep <- setdiff(keep, nzv)
    dropped <- c(dropped, nzv)
  }

  repeat {
    if (length(keep) == 0L) break

    fm <- as.formula(paste(response, "~", paste(keep, collapse = "+")))
    fit_lm <- try(lm(fm, data = train_df), silent = TRUE)
    if (inherits(fit_lm, "try-error")) break

    # 1) Handle aliasing first: drop any predictors with NA coefficients
    aliased <- names(coef(fit_lm))[is.na(coef(fit_lm))]
    aliased <- intersect(aliased, keep)
    if (length(aliased)) {
      keep    <- setdiff(keep, aliased)
      dropped <- c(dropped, aliased)
      next  # re-fit with reduced set
    }

    # 2) Now compute VIF on the estimable set
    v <- try(suppressWarnings(car::vif(fit_lm)), silent = TRUE)
    if (inherits(v, "try-error")) {
      # If something unexpected still happens, bail out of pruning
      break
    }
    if (is.matrix(v)) v <- diag(v)

    vmax <- max(v, na.rm = TRUE)
    if (!is.finite(vmax) || is.nan(vmax) || vmax <= thresh) break

    drop_this <- names(v)[which.max(v)]
    keep    <- setdiff(keep, drop_this)
    dropped <- c(dropped, drop_this)
  }

  list(keep = keep, dropped = unique(dropped))
}

# alias-tolerant VIF (logging helper) 
safe_vif_long <- function(df, response, preds) {
  preds <- unique(preds)
  preds <- preds[!is.na(preds) & preds %in% names(df)]
  if (length(preds) == 0L) {
    return(tibble::tibble(predictor = character(0), vif = numeric(0)))
  }

  fm  <- as.formula(paste(response, "~", paste(preds, collapse = "+")))
  fit <- try(stats::lm(fm, data = df), silent = TRUE)
  if (inherits(fit, "try-error")) {
    return(tibble::tibble(predictor = preds, vif = NA_real_))
  }

  # drop aliased (non-estimable) predictors
  aliased   <- names(coef(fit))[is.na(coef(fit))]
  aliased   <- intersect(aliased, preds)
  estimable <- setdiff(preds, aliased)

  vif_vec <- setNames(rep(NA_real_, length(preds)), preds)

  if (length(estimable) >= 1) {
    if (length(estimable) == 1) {
      vif_est <- setNames(1, estimable)              # VIF = 1 for single predictor
    } else {
      fm_est  <- as.formula(paste(response, "~", paste(estimable, collapse = "+")))
      fit_est <- stats::lm(fm_est, data = df)
      v <- suppressWarnings(car::vif(fit_est))
      if (is.matrix(v)) v <- diag(v)
      vif_est <- setNames(as.numeric(v), names(v))
    }
    vif_vec[names(vif_est)] <- unname(vif_est)
  }

  tibble::tibble(predictor = names(vif_vec), vif = as.numeric(vif_vec))
}

# For each dropped var, find the kept var with the largest |Pearson r|
most_corr_partner_local <- function(df, dropped_vars, kept_vars, response_col, method = "pearson") {
  dropped_vars <- intersect(dropped_vars, names(df))
  kept_vars    <- intersect(kept_vars,    names(df))
  if (length(dropped_vars) == 0 || length(kept_vars) == 0) {
    return(tibble::tibble(predictor = character(), partner = character(), r = numeric()))
  }

  # numeric-only from union (exclude response)
  union_vars <- setdiff(unique(c(dropped_vars, kept_vars)), response_col)
  num_df <- df %>%
    dplyr::select(dplyr::any_of(union_vars)) %>%
    dplyr::mutate(dplyr::across(dplyr::everything(), ~ suppressWarnings(as.numeric(.)))) %>%
    dplyr::select(where(is.numeric))

  # drop <3 finite or zero-variance
  keep_cols <- num_df %>%
    summarise(across(everything(), ~{
      v <- .x[is.finite(.x)]
      (length(v) >= 3) && stats::sd(v) > 0
    })) %>% as.list() %>% unlist()
  num_df <- num_df[, keep_cols, drop = FALSE]
  if (ncol(num_df) == 0) {
    return(tibble::tibble(predictor = character(), partner = character(), r = numeric()))
  }

  left  <- intersect(dropped_vars, colnames(num_df))
  right <- intersect(kept_vars,    colnames(num_df))
  if (length(left) == 0 || length(right) == 0) {
    return(tibble::tibble(predictor = character(), partner = character(), r = numeric()))
  }

  cors <- suppressWarnings(cor(num_df[, left,  drop = FALSE],
                               num_df[, right, drop = FALSE],
                               use = "pairwise.complete.obs",
                               method = method))
  if (!is.matrix(cors)) cors <- as.matrix(cors)
  if (length(left) == 1 && length(right) > 1) cors <- t(cors)

  purrr::map_dfr(seq_along(left), function(i) {
    rvec <- cors[i, ]
    if (length(rvec) == 0 || all(is.na(rvec))) {
      tibble::tibble(predictor = left[i], partner = NA_character_, r = NA_real_)
    } else {
      j <- which.max(abs(rvec))
      tibble::tibble(predictor = left[i], partner = names(rvec)[j], r = unname(rvec[j]))
    }
  })
}



```

*More model setup: AR(1) structure, fixed effects, etc.*
```{r}
# AR(1) correlation structure for gls (single-site) 
make_cor_ar1_local <- function(date_col = "Datetime") {
  corAR1(form = as.formula(paste0("~ as.numeric(", date_col, ")")))
}

# Tidy fixed effects for gls 
tidy_gls_fixed <- function(fit) {
  tt <- summary(fit)$tTable
  tibble(
    term      = rownames(tt),
    estimate  = as.numeric(tt[, "Value"]),
    std.error = as.numeric(tt[, "Std.Error"]),
    statistic = as.numeric(tt[, "t-value"]),
    p.value   = as.numeric(tt[, "p-value"])
  )
}

# Safe gls fit 
fit_gls_safe <- function(fixed_formula, data, date_col = "Datetime") {
  nlme::gls(
    model       = fixed_formula,
    correlation = corAR1(form = as.formula(paste0("~ as.numeric(", date_col, ")"))),
    data        = data,
    method      = "REML",
    na.action   = na.omit,
    control     = nlme::glsControl(
      msMaxIter  = 200,   # optimizer iterations
      msTol      = 1e-7,  # optimizer tolerance (like step tolerance)
      tolerance  = 1e-6,  # convergence tolerance
      msVerbose  = FALSE
    )
  )
}
```

*GLS training fxn*
```{r}
# Core trainer (per-site, per-metric, per-set) 
train_gls_local <- function(data,
                            response,
                            predictors,
                            k = 5,
                            site_col = "Site",
                            date_col = "Datetime",
                            vif_thresh = 5) {

  base_df <- data %>%
    dplyr::select(all_of(c(site_col, date_col, response, predictors))) %>%
    tidyr::drop_na()

  # weekly blocked folds (you already have this helper defined globally)
  folds <- make_weekly_folds(base_df, k = k, date_col = date_col)

  oof_list <- vector("list", k)
  fold_kept_preds <- vector("list", k)
  fold_dropped_preds <- vector("list", k)

  for (i in seq_len(k)) {
    val_idx <- folds$indexOut[[i]]
    trn_idx <- folds$index[[i]]

    train_df <- base_df[trn_idx, , drop = FALSE]
    val_df   <- base_df[val_idx, , drop = FALSE]

    vp <- vif_prune(train_df, response, predictors, thresh = vif_thresh)
    kept <- vp$keep

    fixed_formula <- if (length(kept) > 0) {
      reformulate(termlabels = kept, response = response)
    } else {
      as.formula(paste(response, "~ 1"))
    }

    fit <- fit_gls_safe(fixed_formula, train_df, date_col = date_col)

    val_x <- val_df %>% dplyr::select(all_of(c(site_col, date_col, response, kept))) %>% tidyr::drop_na()
    preds <- as.numeric(predict(fit, newdata = val_x))

    oof_list[[i]] <- tibble(
      Site = val_x[[site_col]],
      Date = val_x[[date_col]],
      obs  = val_x[[response]],
      pred = preds
    )
    fold_kept_preds[[i]]    <- kept
    fold_dropped_preds[[i]] <- vp$dropped
  }

  oof <- bind_rows(oof_list)
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # Fit final on full data with VIF on full set
  vp_full <- vif_prune(base_df, response, predictors, vif_thresh)
  kept_full    <- if (length(vp_full$keep) > 0) vp_full$keep else character(0)
  dropped_full <- vp_full$dropped

  fixed_full <- if (length(kept_full) > 0) {
    reformulate(termlabels = kept_full, response = response)
  } else {
    as.formula(paste(response, "~ 1"))
  }
  fit_full <- fit_gls_safe(fixed_full, base_df, date_col = date_col)

  coefs <- tidy_gls_fixed(fit_full)
  phi <- tryCatch(as.numeric(coef(fit_full$modelStruct$corStruct, unconstrained = FALSE)),
                  error = function(e) NA_real_)

    list(
    response          = response,
    model_type        = "gls_ar1",
    predictors_init   = predictors,
    vif_thresh        = vif_thresh,
    fold_kept         = fold_kept_preds,
    fold_dropped      = fold_dropped_preds,
    kept_full         = kept_full,
    dropped_full      = dropped_full,
    metrics_overall   = mets$overall,
    metrics_by_site   = mets$by_site,
    oof_predictions   = oof,
    fit_object        = fit_full,
    fixed_effects     = coefs,
    ar1_phi           = phi,
    full_data         = base_df   
  )

}

# Runner over local files (per-site, both sets A/B)
run_all_lme_local <- function(data_dir,
                              k = 5,
                              vif_thresh = 5) {
  files <- list.files(data_dir, pattern = "_(max|mean|range)_WT_model_data\\.csv$", full.names = TRUE)
  results <- list()

  for (fp in files) {
    bn <- basename(fp)
    m <- regexec("^([A-Za-z0-9._-]+)_(max|mean|range)_WT_model_data\\.csv$", bn)
    parts <- regmatches(bn, m)[[1]]
    if (length(parts) != 3) next
    site   <- parts[2]
    metric <- parts[3]

    message(">>> ", site, " | ", metric, "_WT")

    df <- read.csv(fp, stringsAsFactors = FALSE) %>% fix_dates()

    # Response + predictors for Set A/B
    response_col <- resolve_response_col(df, metric)
    # Set A: all predictors
    preds_A <- get_predictors_local(df, response_col, mode = "all")
    # Set B: remote-only
    preds_B <- get_predictors_local(df, response_col, mode = "remote")

    # Fit A
    if (length(preds_A) > 0) {
      keyA <- paste(site, metric, "A_gls", sep = "_")
      results[[keyA]] <- train_gls_local(
        data = df, response = response_col, predictors = preds_A,
        k = k, site_col = "Site", date_col = "Datetime", vif_thresh = vif_thresh
      )
    } else {
      warning("No predictors for ", site, " ", metric, " Set A; skipping.")
    }

    # Fit B
    if (length(preds_B) > 0) {
      keyB <- paste(site, metric, "B_gls", sep = "_")
      results[[keyB]] <- train_gls_local(
        data = df, response = response_col, predictors = preds_B,
        k = k, site_col = "Site", date_col = "Datetime", vif_thresh = vif_thresh
      )
    } else {
      warning("No predictors for ", site, " ", metric, " Set B; skipping.")
    }
  }

  results
}

```

*Running GLS model and saving outputs*
```{r}
# Paths
data_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/local_mod_data"
out_dir  <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_lme_results"
dir.create(out_dir, recursive = TRUE, showWarnings = FALSE)

# Run
local_lme_results <- run_all_lme_local(data_dir, k = 5, vif_thresh = 5)

# Save full object
saveRDS(local_lme_results, file.path(out_dir, "local_lme_results.rds"))


# ONE consolidated metrics table (per Site × Metric × Set)
metrics_local_gls <- purrr::imap_dfr(local_lme_results, function(res, nm){
  parts <- strsplit(nm, "_", fixed = TRUE)[[1]]
  Site <- parts[1]; Metric <- parts[2]; Set <- parts[3]; alg <- "gls_ar1"

  tibble::tibble(
    Site, Metric, Set, alg,
    RMSE = res$metrics_overall$RMSE,
    MAE  = res$metrics_overall$MAE,
    R2   = res$metrics_overall$R2,
    NSE  = res$metrics_overall$NSE,
    k            = length(res$fold_kept),        # number of folds used
    n_oof        = nrow(res$oof_predictions),    # out-of-fold predictions count
    n_pred_init  = length(res$predictors_init),  # candidates before VIF
    n_pred_kept  = length(res$kept_full),        # kept after VIF pruning
    ar1_phi      = res$ar1_phi                   # AR(1) phi
  )
})

readr::write_csv(metrics_local_gls, file.path(out_dir, "gls_metrics_local.csv"))


# Fixed effects & AR(1) phi 
fixed_effects_local_lme <- purrr::imap_dfr(local_lme_results, ~{
  .x$fixed_effects %>% dplyr::mutate(model_id = .y, ar1_phi = .x$ar1_phi)
}, .id = NULL) %>%
  tidyr::separate(model_id, into = c("Site","Metric","Set","alg"), sep = "_") %>%
  dplyr::mutate(alg = "gls_ar1") %>%
  dplyr::relocate(Site, Metric, Set, alg, ar1_phi)

readr::write_csv(fixed_effects_local_lme, file.path(out_dir, "gls_fixed_effects_local.csv"))


# VIF log (before/after/dropped) — with most-correlated kept partner for DROPPED
vif_log_local_lme <- purrr::imap_dfr(local_lme_results, function(res, model_id) {
  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  Site <- parts[1]; Metric <- parts[2]; Set <- parts[3]

  vb <- safe_vif_long(res$full_data, res$response, res$predictors_init) %>%
        dplyr::mutate(stage = "before")
  va <- safe_vif_long(res$full_data, res$response, res$kept_full) %>%
        dplyr::mutate(stage = "after")
  vd <- vb %>%
        dplyr::filter(predictor %in% res$dropped_full) %>%
        dplyr::mutate(stage = "dropped")

  # Add partner + r for DROPPED vs KEPT
  corr_partners <- most_corr_partner_local(
    df = res$full_data,
    dropped_vars = res$dropped_full,
    kept_vars    = res$kept_full,
    response_col = res$response,
    method = "pearson"
  )

  vd <- vd %>%
    dplyr::left_join(corr_partners, by = "predictor") %>%
    { if (!"partner" %in% names(.)) dplyr::mutate(., partner = NA_character_) else . } %>%
    { if (!"r" %in% names(.))       dplyr::mutate(., r = NA_real_)       else . } %>%
    dplyr::rename(partner_most_corr = partner, r_with_partner = r)

  dplyr::bind_rows(vb, va, vd) %>%
    dplyr::mutate(Site = Site, Metric = Metric, Set = Set, alg = "gls_ar1",
                  dropped = stage == "dropped") %>%
    dplyr::select(Site, Metric, Set, alg, stage, predictor, vif,
                  partner_most_corr, r_with_partner, dropped)
}, .id = NULL)

readr::write_csv(
  vif_log_local_lme,
  file.path(out_dir, "gls_vif_log_long_local.csv")
)


```

*Permutation variable importances*
```{r}
# Permutation VI for ALL FINAL LOCAL MODELS
# (XGB, RFR, GLS/AR1) -> one consolidated CSV
library(vip)

set.seed(42)
options(vip.smaller_is_better = TRUE)

# Paths
DATA_DIR  <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/local_mod_data"
FINAL_CSV <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_final.csv"
ORIG_RDS  <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/local_models_refined.rds"
PASS2_DIR <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results_pass2"
GLS_RDS   <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_lme_results/local_lme_results.rds"
OUT_DIR   <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results"
dir.create(OUT_DIR, recursive = TRUE, showWarnings = FALSE)

# Helpers 
rmse_loss <- function(truth, estimate) sqrt(mean((truth - estimate)^2, na.rm = TRUE))
pred_wrap_caret <- function(object, newdata) predict(object, newdata = newdata)
pred_wrap_gls   <- function(object, newdata) predict(object, newdata = newdata)

# Final model selection (ML rows only)
final_ml <- readr::read_csv(FINAL_CSV, show_col_types = FALSE) %>%
  mutate(Model = toupper(Model)) %>%
  filter(Model %in% c("XGB","RFR"))

# Original ML model list
orig_list <- if (file.exists(ORIG_RDS)) readRDS(ORIG_RDS) else NULL

# Load a final ML model object (orig vs pass2) for a given key
load_final_ml_model <- function(site, metric, setAB, model, source) {
  alg_tag <- tolower(model)  # "xgb" | "rfr"
  source  <- tolower(source)
  if (source == "pass2") {
    fp <- file.path(PASS2_DIR, sprintf("%s_%s_%s_%s_pass2.rds", site, metric, setAB, alg_tag))
    if (!file.exists(fp)) stop("Pass-2 file not found: ", fp)
    readRDS(fp)
  } else {
    if (is.null(orig_list)) stop("Original models list RDS not found: ", ORIG_RDS)
    key <- paste(site, metric, setAB, alg_tag, sep = "_")
    if (!key %in% names(orig_list)) stop("Original model not in list: ", key)
    orig_list[[key]]
  }
}

# Build X/y from the on-disk local CSV, aligned to a given predictor set
build_train_xy <- function(site, metric, predictors) {
  in_fp <- file.path(DATA_DIR, sprintf("%s_%s_WT_model_data.csv", site, metric))
  stopifnot(file.exists(in_fp))
  df <- read.csv(in_fp, stringsAsFactors = FALSE) %>% fix_dates()
  response_col <- resolve_response_col(df, metric)
  # Keep only response + predictors; drop rows with NA
  combo <- df %>% dplyr::select(dplyr::all_of(c(response_col, predictors))) %>% tidyr::drop_na()
  list(
    y = combo[[response_col]],
    X = combo %>% dplyr::select(dplyr::all_of(predictors)),
    response = response_col
  )
}

NSIMS <- 20

# Permutation VI for FINAL ML models (XGB/RFR)
vi_ml_perm <- purrr::pmap_dfr(
  final_ml %>% dplyr::select(Site, Metric, Set, Model, source),
  function(Site, Metric, Set, Model, source) {
    # Load the chosen (orig or pass2) model
    res <- load_final_ml_model(Site, Metric, Set, Model, source)

    # Predictors actually used in that fit
    preds <- res$predictors
    if (is.null(preds) || !length(preds)) {
      warning("No predictors for ", paste(Site, Metric, Set, Model, source, sep=" / "), "; skipping.")
      return(tibble())
    }

    tr <- build_train_xy(Site, Metric, preds)

    vi <- vip::vi_permute(
      object            = res$fit_object,   # caret::train
      train             = tr$X,
      target            = tr$y,
      metric            = rmse_loss,
      pred_wrapper      = pred_wrap_caret,
      nsim              = NSIMS,
      keep              = TRUE,
      parallel          = FALSE,
      smaller_is_better = TRUE
    )

    tibble(
      Site = Site, Metric = Metric, Set = Set,
      Model = toupper(Model), source = tolower(source),
      Variable   = vi$Variable,
      Importance = vi$Importance,
      Stderr     = vi$StDev,
      nsim       = NSIMS,
      type       = "permutation"
    )
  }
)


# Permutation VI for GLS (AR1) local models
vi_gls_perm <- {
  if (file.exists(GLS_RDS)) {
    gls_list <- readRDS(GLS_RDS)
    purrr::imap_dfr(gls_list, function(res, key){
      parts <- strsplit(key, "_", fixed = TRUE)[[1]]
      Site <- parts[1]; Metric <- parts[2]; Set <- parts[3]
      kept <- res$kept_full
      if (is.null(kept) || !length(kept)) return(tibble())

      # Use the stored full_data to avoid any mismatch
      df <- res$full_data
      response <- res$response

      combo <- df %>% dplyr::select(dplyr::all_of(c(response, kept))) %>% tidyr::drop_na()
      y <- combo[[response]]
      X <- combo %>% dplyr::select(dplyr::all_of(kept))

      vi <- vip::vi_permute(
        object            = res$fit_object,   # nlme::gls
        train             = X,
        target            = y,
        metric            = rmse_loss,
        pred_wrapper      = pred_wrap_gls,
        nsim              = NSIMS,
        keep              = TRUE,
        parallel          = FALSE,
        smaller_is_better = TRUE
      )

      tibble(
        Site = Site, Metric = Metric, Set = Set,
        Model = "GLS", source = "gls",
        Variable   = vi$Variable,
        Importance = vi$Importance,
        Stderr     = vi$StDev,
        nsim       = NSIMS,
        type       = "permutation"
      )
    })
  } else {
    tibble()
  }
}


# Consolidate (ALL models) and write ONE file
varimp_perm_local_ALL_final <- bind_rows(vi_ml_perm, vi_gls_perm) %>%
  arrange(Site, Metric, Set, Model, desc(Importance))

OUT_FILE <- file.path(OUT_DIR, "varimp_perm_local_ALL_final.csv")
readr::write_csv(varimp_perm_local_ALL_final, OUT_FILE)

message("Wrote: ", OUT_FILE,
        "\nRows: ", nrow(varimp_perm_local_ALL_final))
```

*Variable importance plots*
```{r}
library(forcats)
library(stringr)
library(ggh4x)

vi_file    <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results/local_results/varimp_perm_local_ALL_final.csv"  # <-- set to your file
output_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/figures"

# Load & prep
vi_raw <- readr::read_csv(vi_file, show_col_types = FALSE)

vi <- vi_raw %>%
  filter(Set == "A") %>%
  mutate(
    alg = dplyr::case_when(
      Model == "GLS" ~ "GLS",
      Model == "RFR" ~ "RFR",
      Model == "XGB" ~ "XGB",
      TRUE ~ Model
    ),
    response = dplyr::case_when(
      Metric == "max"   ~ "Max",
      Metric == "mean"  ~ "Mean",
      Metric == "range" ~ "Range",
      TRUE ~ Metric
    ),
    alg = factor(alg, levels = c("GLS","RFR","XGB")),
    response = factor(response, levels = c("Max","Mean","Range"))
  ) %>%
  select(Site, response, alg, Variable, Importance, Stderr)

# SITE-SPECIFIC mini-global plot 
make_site_plot <- function(site_name, top_k = 12, clamp_LME = FALSE) {

  # 1) choose top-k variables *within this site x response* (mean across algs)
  site_top <- vi %>%
    filter(Site == site_name) %>%
    group_by(response, Variable) %>%
    summarise(mean_imp = mean(Importance, na.rm = TRUE), .groups = "drop") %>%
    filter(!is.na(mean_imp)) %>%
    group_by(response) %>%
    slice_max(order_by = mean_imp, n = top_k, with_ties = FALSE) %>%
    arrange(response, desc(mean_imp)) %>%
    mutate(var_key = paste0(response, "__", Variable)) %>%
    ungroup()

  # 2) build plotting frame for this site using its own top-k order
  plot_df <- vi %>%
    filter(Site == site_name) %>%
    mutate(
      var_key = paste0(response, "__", Variable),
      Importance = dplyr::coalesce(Importance, 0)  # show 0 if a model didn't produce VI
    ) %>%
    inner_join(site_top, by = c("response","Variable","var_key"))

  p <- ggplot(plot_df, aes(x = forcats::fct_reorder(var_key, mean_imp), y = Importance)) +
    geom_col(width = 0.75) +
    geom_errorbar(aes(ymin = pmax(0, Importance - Stderr),
                      ymax = Importance + Stderr),
                  width = 0.2, alpha = 0.5) +
    geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
    coord_flip() +
    facet_grid(response ~ alg, scales = "free_y", space = "free_y") +
    scale_x_discrete(labels = function(x) sub("^.*__", "", x)) +
    labs(
      title = paste0("Local variable importance — ", site_name),
      x = "Predictor",
      y = "Increase in RMSE (°C)"
    ) +
    theme_minimal(base_size = 11) +
    theme(
      panel.grid.minor = element_blank(),
      strip.text = element_text(face = "bold"),
      plot.title = element_text(face = "bold"),
      axis.title.y = element_text(margin = margin(r = 10)),
      panel.spacing = unit(0.5, "lines"),
      plot.margin   = margin(10, 10, 10, 10)
    )

  if (clamp_LME) {
    p <- p + ggh4x::facetted_pos_scales(
      y = list(alg == "LME" ~ scale_y_continuous(limits = c(0, 1)))
    )
  }
  p
}

# make & save all sites
sites <- vi %>% distinct(Site) %>% arrange(Site) %>% pull(Site)

purrr::walk(sites, function(s) {
  p <- make_site_plot(s, top_k = 12, clamp_LME = FALSE)
  ggsave(
    filename = file.path("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/figures/varimp_local",
                         paste0("varimp_local_", gsub("\\s+", "_", s), ".png")),
    plot = p, width = 8.5, height = 9, units = "in", dpi = 300, bg = "white"
  )
})

```

*Local obs vs pred plots*
```{r}
# Labels
resp_lab <- c(max = "Max WT", mean = "Mean WT", range = "Range WT")
alg_lab  <- c(lme = "GLS",    rfr  = "RFR",     xgb   = "XGB")

# 1) Collect Set A OOF predictions from your local objects
collect_local_oof_setA <- function(local_results_refined, local_lme_results) {
  # RFR/XGB
  df_ml <- purrr::imap_dfr(local_results_refined, function(x, nm){
    parts <- strsplit(nm, "_", fixed = TRUE)[[1]]  # <Site>_<metric>_<Set>_<model>
    if (length(parts) != 4L || parts[3] != "A") return(NULL)
    tibble(
      Site     = parts[1],
      response = parts[2],       # max / mean / range
      set      = parts[3],       # A
      alg      = parts[4],       # rfr / xgb
      Date     = x$oof_predictions$Date,
      obs      = x$oof_predictions$obs,
      pred     = x$oof_predictions$pred
    )
  })

  # GLS (normalize to alg="lme")
  df_gls <- purrr::imap_dfr(local_lme_results, function(x, nm){
    parts <- strsplit(nm, "_", fixed = TRUE)[[1]]  # <Site>_<metric>_<Set>_gls
    if (length(parts) != 4L || parts[3] != "A") return(NULL)
    tibble(
      Site     = parts[1],
      response = parts[2],
      set      = parts[3],
      alg      = "lme",
      Date     = x$oof_predictions$Date,
      obs      = x$oof_predictions$obs,
      pred     = x$oof_predictions$pred
    )
  })

  bind_rows(df_ml, df_gls) %>%
    mutate(
      response = factor(response, levels = c("max","mean","range"), labels = resp_lab),
      alg      = factor(alg,      levels = c("lme","rfr","xgb"),    labels = alg_lab)
    ) %>%
    arrange(response, alg, Site, Date)
}

# Build OOF df (requires local_results_refined + local_lme_results in env)
oof_local_A <- collect_local_oof_setA(local_results_refined, local_lme_results)

# 2) Helpers for scales/metrics
int_breaks <- function(lims, target = 6L) {
  lo <- floor(lims[1]); hi <- ceiling(lims[2])
  span <- max(1, hi - lo)
  step <- max(1L, round(span / target))
  seq(lo, hi, by = step)
}

limits_by_response <- function(df) {
  df %>%
    group_by(response) %>%
    summarise(
      lo = floor(min(c(obs, pred), na.rm = TRUE)),
      hi = ceiling(max(c(obs, pred), na.rm = TRUE)),
      .groups = "drop"
    )
}

# 3) Main plotting function: one figure per response (rows=models, cols=sites)
make_response_grid <- function(dat, resp_label, site_order = NULL, show_metrics = TRUE) {
  dd <- dat %>% filter(response == resp_label)

  # order sites (alphabetical by default)
  if (is.null(site_order)) {
    site_order <- dd %>% distinct(Site) %>% arrange(Site) %>% pull(Site)
  }

  # common limits/breaks for this response
  lim_tbl <- limits_by_response(dd)
  lim_row <- lim_tbl %>% filter(response == resp_label)
  lims <- c(lim_row$lo, lim_row$hi)
  brk  <- int_breaks(lims, target = 6L)

  # per-panel metrics (Site x alg for this response)
  mets <- dd %>%
    group_by(Site, alg) %>%
    tidyr::nest() %>%
    mutate(
      slope = purrr::map_dbl(data, ~ coef(lm(pred ~ obs, data = .x))[2]),
      rmse  = purrr::map_dbl(data, ~ sqrt(mean((.x$pred - .x$obs)^2, na.rm = TRUE)))
    ) %>%
    select(-data)

  # annotation position (top-left of each panel)
  ann_y <- lims[2] - (diff(lims) * 0.05)
  ann_x <- lims[1] + (diff(lims) * 0.03)

  ggplot(
    dd %>% mutate(Site = factor(Site, levels = site_order)),
    aes(obs, pred)
  ) +
    geom_point(alpha = 0.65, size = 1.6, stroke = 0.25) +
    geom_abline(slope = 1, intercept = 0, linetype = "dashed", linewidth = 0.7) +
    geom_smooth(method = "lm", se = FALSE, linewidth = 0.9) +
    { if (show_metrics)
        geom_text(
          data = mets %>% mutate(Site = factor(Site, levels = site_order)),
          aes(x = ann_x, y = ann_y,
              label = sprintf("slope=%.2f\nRMSE=%.2f", slope, rmse)),
          hjust = 0, vjust = 1, size = 3, inherit.aes = FALSE
        )
    } +
    facet_grid(rows = vars(alg), cols = vars(Site)) +
    coord_equal(ratio = 1) +
    scale_x_continuous(limits = lims, breaks = brk, minor_breaks = NULL,
                       labels = scales::label_number(accuracy = 1)) +
    scale_y_continuous(limits = lims, breaks = brk, minor_breaks = NULL,
                       labels = scales::label_number(accuracy = 1)) +
    labs(
      title = paste0("Observed vs Predicted — ", resp_label),
      x = "Observed (°C)",
      y = "Predicted (°C)"
    ) +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.35),
      strip.text = element_text(face = "bold"),
      legend.position = "none",
      panel.spacing = unit(6, "pt"),
      plot.margin = unit(c(5, 6, 5, 6), "pt")
    )
}

# 4) Make & save three per-response grids
out_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/figures/obs_pred_local"
if (!dir.exists(out_dir)) dir.create(out_dir, recursive = TRUE)

responses <- c("Max WT","Mean WT","Range WT")
purrr::walk(responses, function(rsp) {
  p <- make_response_grid(oof_local_A, rsp, show_metrics = TRUE)
  ggsave(
    filename = file.path(out_dir, paste0("obs_vs_pred_local_", gsub("\\s+","_", rsp), ".png")),
    plot = p, width = 12, height = 8.5, units = "in", dpi = 300, bg = "white"
  )
})

```



