---
title: "Global wetland ML models"
author: "Michael Krochta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Set up**
```{r}
library(dplyr)
library(lubridate)
library(caret)
library(ranger)     
library(xgboost)    
library(tibble)
library(stringr)
library(tidyr)
library(purrr)
library(doParallel)
set.seed(42)        # Reproducibility for folds/shuffles
```

**Loading data**
```{r}
max_df  <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/max_global_data.csv",  stringsAsFactors = FALSE)
mean_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/mean_global_data.csv", stringsAsFactors = FALSE)
range_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/range_global_data.csv",stringsAsFactors = FALSE)

# Ensure consistent column names and types
# - Datetime to Date
fix_dates <- function(df) {
  if (!inherits(df$Datetime, "Date")) {
    df$Datetime <- as.Date(df$Datetime, tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y"))
  }
  df
}
max_df   <- fix_dates(max_df)
mean_df  <- fix_dates(mean_df)
range_df <- fix_dates(range_df)

# Response column names (adjust if different)
resp_cols <- list(
  max   = "max_WT",
  mean  = "mean_WT",
  range = "range_WT"
)
```

*Define "Remote-only" exclusion list (field-derived vars)*
```{r}
remote_exclude_exact <- c(
  "WL_daily_mean","WL_daily_max","WL_daily_min","WL_daily_range",
  "WL_mean_2day","WL_mean_3day","WL_mean_7day",
  "WL_max_2day","WL_max_3day","WL_max_7day",
  "WL_min_2day","WL_min_3day","WL_min_7day",
  "GWT_mean","GWT_max","GWT_min","GWT_range",
  "GWT_mean_2day","GWT_mean_3day","GWT_mean_7day",
  "GWT_max_2day","GWT_max_3day","GWT_max_7day",
  "GWT_min_2day","GWT_min_3day","GWT_min_7day",
  "GWL_mean","GWL_max","GWL_min","GWL_range",
  "GWL_mean_2day","GWL_mean_3day","GWL_mean_7day",
  "GWL_max_2day","GWL_max_3day","GWL_max_7day",
  "GWL_min_2day","GWL_min_3day","GWL_min_7day", "Water_area_m2_3d"
)
# Also remove any columns that *start with* these prefixes
remote_exclude_prefix <- c("WL_", "GWT_", "GWL_", "Water")

# Build predictor vector for a given df + mode
get_predictors <- function(df, response_col, mode = c("all", "remote")) {
  mode <- match.arg(mode)
  base <- setdiff(names(df), c(response_col, "Site", "Datetime")) # never model on ID/time
  if (mode == "all") return(base)
  # Remote-only: drop exact matches + prefixed columns if present
  drop_prefixed <- unlist(lapply(remote_exclude_prefix, function(p) grep(paste0("^", p), base, value = TRUE)))
  setdiff(base, unique(c(remote_exclude_exact, drop_prefixed)))
}
```


**CV: Build blocked folds by week**
*Keeps whole weeks together (reduces inflated accuracy from adjacent days)*
```{r}
make_weekly_folds <- function(df, k = 5, date_col = "Datetime") {
  df <- df %>% mutate(.week = floor_date(.data[[date_col]], unit = "week", week_start = 1))
  weeks <- unique(df$.week)
  set.seed(42)
  weeks <- sample(weeks)        # shuffle weeks
  folds <- split(weeks, cut(seq_along(weeks), breaks = k, labels = FALSE))

  index <- list(); indexOut <- list()
  for (i in seq_len(k)) {
    val_weeks <- folds[[i]]
    val_idx <- which(df$.week %in% val_weeks)
    trn_idx <- setdiff(seq_len(nrow(df)), val_idx)
    index[[paste0("Fold", i)]]    <- trn_idx
    indexOut[[paste0("Fold", i)]] <- val_idx
  }
  list(index = index, indexOut = indexOut)
}
```

**Metrics**
```{r}
nse <- function(obs, pred) {
  1 - sum((obs - pred)^2, na.rm = TRUE) / sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
}

# Compute overall + by-site metrics from out-of-fold predictions
compute_metrics <- function(pred_df, truth_col = "obs", pred_col = "pred", site_col = "Site") {
  # Overall
  overall <- tibble(
    RMSE = RMSE(pred_df[[pred_col]], pred_df[[truth_col]]),
    MAE  = MAE (pred_df[[pred_col]], pred_df[[truth_col]]),
    R2   = R2  (pred_df[[pred_col]], pred_df[[truth_col]]),
    NSE  = nse (pred_df[[truth_col]], pred_df[[pred_col]])
  )

  # By-site
  by_site <- pred_df %>%
    group_by(.data[[site_col]]) %>%
    summarise(
      RMSE = RMSE(.data[[pred_col]], .data[[truth_col]]),
      MAE  = MAE (.data[[pred_col]], .data[[truth_col]]),
      R2   = R2  (.data[[pred_col]], .data[[truth_col]]),
      NSE  = nse (.data[[truth_col]], .data[[pred_col]]),
      .groups = "drop"
    ) %>%
    rename(Site = .data[[site_col]])

  list(overall = overall, by_site = by_site)
}
```

**Hyperparameter grids (Stage 2: refined)**
```{r}
rf_grid_refined <- function(p) {
  expand.grid(
    mtry          = sort(unique(pmax(1, c(
                        floor(sqrt(p))-1, floor(sqrt(p)), floor(sqrt(p))+1,
                        floor(p/4), floor(p/3), floor(p/2), ceiling(2*p/3)
                      )))),
    splitrule     = "variance",
    min.node.size = c(1, 2, 3, 5, 7, 10, 15),
    stringsAsFactors = FALSE
  )
}

xgb_grid_refined <- expand.grid(
  nrounds          = c(400, 800),
  eta              = c(0.05, 0.1),
  max_depth        = c(6, 9, 12),
  min_child_weight = c(1, 4),
  subsample        = c(0.8, 0.9),
  colsample_bytree = c(0.8, 1.0),
  gamma            = c(0, 0.1, 1.0)
)

```

**Core model trainer (XGB or RFR) with blocked CV + saved OOF predictions**
```{r}
train_model <- function(data,
                        response,
                        predictors,
                        model_type = c("xgb", "rfr"),
                        k = 5,
                        site_col = "Site",
                        date_col = "Datetime",
                        tune_grid = NULL) {

  model_type <- match.arg(model_type)

  # 1) Keep only needed columns; drop rows with NA in response/predictors
  df <- data %>% select(all_of(c(site_col, date_col, response, predictors))) %>% na.omit()

  # 2) Build blocked folds (weekly)
  folds <- make_weekly_folds(df, k = k, date_col = date_col)

  # 3) caret control: use our folds + save OOF preds for by-site metrics
  ctrl <- trainControl(
  method          = "cv",
  number          = k,
  index           = folds$index,
  indexOut        = folds$indexOut,
  verboseIter     = FALSE,
  savePredictions = "final",
  allowParallel   = TRUE      
)

  # 4) Default tuning grids if not provided
  if (is.null(tune_grid)) {
    if (model_type == "rfr") {
      tune_grid <- rf_grid_refined(length(predictors))
    } else {
      tune_grid <- xgb_grid_refined
    }
  }

  # 5) Train
  form <- as.formula(paste(response, "~ ."))
  train_df <- df %>% select(all_of(c(response, predictors)))   # caret formula uses only predictors/response

  if (model_type == "rfr") {
  if (is.null(tune_grid)) tune_grid <- rf_grid_refined(length(predictors))
    fit <- train(
  form,
  data       = train_df,
  method     = "ranger",
  trControl  = ctrl,
  tuneGrid   = tune_grid,
  metric     = "RMSE",
  importance = "permutation",
  num.trees  = 500,
  num.threads = 1            
)
  } else {
    fit <- train(
  form,
  data      = train_df,
  method    = "xgbTree",
  trControl = ctrl,
  tuneGrid  = tune_grid,
  metric    = "RMSE",
  nthread   = 1            
)
  }

  # 6) Out-of-fold predictions joined back to Site for by-site metrics
  # caret's $pred stores CV preds; need the matching rows' Site values
 # --- OOF predictions: align to original rows via caret's rowIndex
if (!"rowIndex" %in% names(fit$pred)) {
  stop("caret did not return 'rowIndex' in fit$pred; cannot align OOF predictions to Site/Date.")
}

oof <- fit$pred %>%
  mutate(
    Site = df[[site_col]][rowIndex],
    Date = df[[date_col]][rowIndex]
  ) %>%
  transmute(Site, Date, obs = obs, pred = pred)

  # 7) Metrics
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # 8) Variable importance
  vimp <- varImp(fit)$importance %>%
    rownames_to_column("Variable") %>%
    arrange(desc(Overall))

  list(
    response     = response,
    model_type   = model_type,
    predictors   = predictors,
    best_params  = fit$bestTune,
    cv_results   = fit$results,
    metrics_overall = mets$overall,
    metrics_by_site = mets$by_site,
    var_importance  = vimp,
    oof_predictions = oof,
    fit_object      = fit
  )
}
```

**Build datasets for each predictor set**
*- Set A (All): exclude Hideaway & Cooper*
*- Set B (Remote-only): include all 7 sites; drop WL/GWT/GWL vars*
```{r}
sites_exclude_all <- c("Hideaway","Cooper")

prep_dataset <- function(df, response_key, mode = c("all","remote")) {
  mode <- match.arg(mode)
  response_col <- resp_cols[[response_key]]

  df2 <- df
  if (mode == "all") {
    df2 <- df2 %>% filter(!(Site %in% sites_exclude_all))
  }
  preds <- get_predictors(df2, response_col, mode = mode)
  list(data = df2, response = response_col, predictors = preds)
}
```

**Run global models for all responses and both predictor sets**
*(XGB and RFR; 5-fold blocked CV)*
```{r}
run_all <- function() {
  results <- list()

  for (resp_key in c("max","mean","range")) {
    # Select the right data frame for this response
    df <- switch(resp_key,
                 max   = max_df,
                 mean  = mean_df,
                 range = range_df)

    # --- Set A: All predictors (5 sites)
    A <- prep_dataset(df, resp_key, mode = "all")
    # XGB
    results[[paste(resp_key,"A_xgb",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"A_rfr",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "rfr", k = 5
    )

    # --- Set B: Remote-only (7 sites)
    B <- prep_dataset(df, resp_key, mode = "remote")
    # XGB
    results[[paste(resp_key,"B_xgb",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"B_rfr",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "rfr", k = 5
    )
  }

  results
}
```

***RUN THE MODELS***
```{r}

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

results_refined <- run_all()

stopCluster(cl)
```


**Evaluating results**
```{r}
names(results_refined)

# Overall metrics
metrics_overall <- imap_dfr(results_refined, ~ .x$metrics_overall, .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg)

# By-site metrics (with OOF counts)
metrics_by_site <- imap_dfr(results_refined, function(x, nm){
  n_by_site <- x$oof_predictions %>% count(Site, name = "n_oof")
  x$metrics_by_site %>%
    left_join(n_by_site, by = "Site") %>%
    mutate(model_id = nm)
}, .id = NULL) %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg,Site)

# Best params in tidy long form (works for both algs)
best_params_long <- imap_dfr(results_refined, ~{
  tibble::as_tibble(.x$best_params) %>% mutate(model_id = .y)
}, .id = NULL) %>%
  pivot_longer(
    -model_id,
    names_to = "param",
    values_to = "value",
    values_ptypes    = list(value = character()),   # <- force char
    values_transform = list(value = as.character)
  ) %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response, set, alg, param, value)

# “Edge of grid?” check (do we need a second pass?)
edge_check <- imap_dfr(results_refined, function(x, nm){
  bt <- x$best_params
  grid <- x$cv_results
  if (is.null(bt) || is.null(grid) || nrow(grid) == 0) {
    return(tibble(model_id = nm, param = NA_character_, value = NA_character_, at_edge = NA))
  }
  ps <- intersect(names(bt), names(grid))
  purrr::map_dfr(ps, function(p){
    col <- grid[[p]]
    if (is.factor(col)) col <- as.character(col)
    val <- bt[[p]]
    at_edge <- if (is.numeric(col)) {
      suppressWarnings(as.numeric(val)) %in% range(col, na.rm = TRUE)
    } else {
      u <- sort(unique(as.character(col)))
      if (length(u) <= 1) FALSE else as.character(val) %in% c(u[1], u[length(u)])
    }
    tibble(model_id = nm, param = p, value = as.character(val), at_edge = at_edge)
  })
}) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response, set, alg, param, value, at_edge)

# Now check after filtering
edge_hits <- edge_check %>% filter(isTRUE(at_edge))

if (nrow(edge_hits) == 0) {
  message("✅ No parameters hit grid edges.")
} else {
  message("⚠️ Some parameters hit grid edges.")
}

edge_hits


# Top variables (internal importance)
top_vars <- function(x, k=10) x$var_importance %>% dplyr::slice_max(Overall, n=k)
top_vars(results_refined$max_A_rfr, 10)
top_vars(results_refined$max_A_xgb, 10)

output_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results"
readr::write_csv(metrics_overall, file.path(output_dir, "metrics_overall.csv"))
readr::write_csv(metrics_by_site, file.path(output_dir, "metrics_by_site.csv"))
readr::write_csv(best_params_long,    file.path(output_dir, "best_params_with_cv_metrics.csv"))

```

*k-fold diagnostic*
```{r}
# Fold-level RMSE variability for each model
fold_disp <- purrr::imap_dfr(results_refined, ~{
  res <- .x$fit_object$pred %>%
    group_by(Resample) %>%
    summarise(RMSE = caret::RMSE(pred, obs), .groups = "drop")
  res$model_id <- .y
  res
}) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  group_by(response, set, alg) %>%
  summarise(
    mean_RMSE = mean(RMSE),
    sd_RMSE   = sd(RMSE),
    cv        = sd_RMSE / mean_RMSE,  # coefficient of variation
    .groups = "drop"
  ) %>%
  arrange(desc(cv))

fold_disp
#These look very stable — all CVs are well below 0.15, which means k = 5 is giving consistent fold-level RMSE across all models.
```

*variable importance (for possibly reconsidering including some variables)*
```{r}
get_top_importance <- function(model_list, top_n = NULL) {
  purrr::imap_dfr(model_list, ~{
    vimp <- .x$var_importance %>%
      mutate(model_id = .y)
  })
}

all_vimp <- get_top_importance(results_refined)

# Rank importance within each model
all_vimp <- all_vimp %>%
  group_by(model_id) %>%
  mutate(rank = rank(-Overall, ties.method = "min")) %>%
  ungroup()

# Predictors consistently ranked low
low_ranked <- all_vimp %>%
  group_by(Variable) %>%
  summarise(mean_rank = mean(rank), max_rank = max(rank), .groups = "drop") %>%
  arrange(desc(mean_rank))

low_ranked

# since the models are currently stable, i'm going to just keep all variables for now.

```
