---
title: "Global wetland ML models"
author: "Michael Krochta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Set up**
```{r}
library(dplyr)
library(lubridate)
library(caret)
library(ranger)     
library(xgboost)    
library(tibble)
library(stringr)
library(tidyr)
library(purrr)
library(doParallel)
library(cowplot)
library(scales)
library(grid) 
library(patchwork)

set.seed(42)        # Reproducibility for folds/shuffles
```

**Loading data**
```{r}
max_df  <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/max_global_data.csv",  stringsAsFactors = FALSE)
mean_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/mean_global_data.csv", stringsAsFactors = FALSE)
range_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/range_global_data.csv",stringsAsFactors = FALSE)

# Ensure consistent column names and types
# - Datetime to Date
fix_dates <- function(df) {
  if (!inherits(df$Datetime, "Date")) {
    df$Datetime <- as.Date(df$Datetime, tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y"))
  }
  df
}
max_df   <- fix_dates(max_df)
mean_df  <- fix_dates(mean_df)
range_df <- fix_dates(range_df)

# Response column names (adjust if different)
resp_cols <- list(
  max   = "max_WT",
  mean  = "mean_WT",
  range = "range_WT"
)
```

*Define "Remote-only" exclusion list (field-derived vars)*
```{r}
remote_exclude_exact <- c(
  "WL_daily_mean","WL_daily_max","WL_daily_min","WL_daily_range",
  "WL_mean_2day","WL_mean_3day","WL_mean_7day",
  "WL_max_2day","WL_max_3day","WL_max_7day",
  "WL_min_2day","WL_min_3day","WL_min_7day",
  "GWT_mean","GWT_max","GWT_min","GWT_range",
  "GWT_mean_2day","GWT_mean_3day","GWT_mean_7day",
  "GWT_max_2day","GWT_max_3day","GWT_max_7day",
  "GWT_min_2day","GWT_min_3day","GWT_min_7day",
  "GWL_mean","GWL_max","GWL_min","GWL_range",
  "GWL_mean_2day","GWL_mean_3day","GWL_mean_7day",
  "GWL_max_2day","GWL_max_3day","GWL_max_7day",
  "GWL_min_2day","GWL_min_3day","GWL_min_7day", "Water_area_m2_3d"
)
# Also remove any columns that *start with* these prefixes
remote_exclude_prefix <- c("WL_", "GWT_", "GWL_", "Water")

# Build predictor vector for a given df + mode
get_predictors <- function(df, response_col, mode = c("all", "remote")) {
  mode <- match.arg(mode)
  base <- setdiff(names(df), c(response_col, "Site", "Datetime")) # never model on ID/time
  if (mode == "all") return(base)
  # Remote-only: drop exact matches + prefixed columns if present
  drop_prefixed <- unlist(lapply(remote_exclude_prefix, function(p) grep(paste0("^", p), base, value = TRUE)))
  setdiff(base, unique(c(remote_exclude_exact, drop_prefixed)))
}
```


**CV: Build blocked folds by week**
*Keeps whole weeks together (reduces inflated accuracy from adjacent days)*
```{r}
make_weekly_folds <- function(df, k = 5, date_col = "Datetime") {
  df <- df %>% mutate(.week = floor_date(.data[[date_col]], unit = "week", week_start = 1))
  weeks <- unique(df$.week)
  set.seed(42)
  weeks <- sample(weeks)        # shuffle weeks
  folds <- split(weeks, cut(seq_along(weeks), breaks = k, labels = FALSE))

  index <- list(); indexOut <- list()
  for (i in seq_len(k)) {
    val_weeks <- folds[[i]]
    val_idx <- which(df$.week %in% val_weeks)
    trn_idx <- setdiff(seq_len(nrow(df)), val_idx)
    index[[paste0("Fold", i)]]    <- trn_idx
    indexOut[[paste0("Fold", i)]] <- val_idx
  }
  list(index = index, indexOut = indexOut)
}
```

**Metrics**
```{r}
nse <- function(obs, pred) {
  1 - sum((obs - pred)^2, na.rm = TRUE) / sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
}

# Compute overall + by-site metrics from out-of-fold predictions
compute_metrics <- function(pred_df, truth_col = "obs", pred_col = "pred", site_col = "Site") {
  # Overall
  overall <- tibble(
    RMSE = RMSE(pred_df[[pred_col]], pred_df[[truth_col]]),
    MAE  = MAE (pred_df[[pred_col]], pred_df[[truth_col]]),
    R2   = R2  (pred_df[[pred_col]], pred_df[[truth_col]]),
    NSE  = nse (pred_df[[truth_col]], pred_df[[pred_col]])
  )

  # By-site
  by_site <- pred_df %>%
    group_by(.data[[site_col]]) %>%
    summarise(
      RMSE = RMSE(.data[[pred_col]], .data[[truth_col]]),
      MAE  = MAE (.data[[pred_col]], .data[[truth_col]]),
      R2   = R2  (.data[[pred_col]], .data[[truth_col]]),
      NSE  = nse (.data[[truth_col]], .data[[pred_col]]),
      .groups = "drop"
    ) %>%
    rename(Site = .data[[site_col]])

  list(overall = overall, by_site = by_site)
}
```

**Hyperparameter grids (Stage 2: refined)**
```{r}
rf_grid_refined <- function(p) {
  expand.grid(
    mtry          = sort(unique(pmax(1, c(
                        floor(sqrt(p))-1, floor(sqrt(p)), floor(sqrt(p))+1,
                        floor(p/4), floor(p/3), floor(p/2), ceiling(2*p/3)
                      )))),
    splitrule     = "variance",
    min.node.size = c(1, 2, 3, 5, 7, 10, 15),
    stringsAsFactors = FALSE
  )
}

xgb_grid_refined <- expand.grid(
  nrounds          = c(400, 800),
  eta              = c(0.05, 0.1),
  max_depth        = c(6, 9, 12),
  min_child_weight = c(1, 4),
  subsample        = c(0.8, 0.9),
  colsample_bytree = c(0.8, 1.0),
  gamma            = c(0, 0.1, 1.0)
)

```

**Core model trainer (XGB or RFR) with blocked CV + saved OOF predictions**
```{r}
train_model <- function(data,
                        response,
                        predictors,
                        model_type = c("xgb", "rfr"),
                        k = 5,
                        site_col = "Site",
                        date_col = "Datetime",
                        tune_grid = NULL) {

  model_type <- match.arg(model_type)

  # 1) Keep only needed columns; drop rows with NA in response/predictors
  df <- data %>% select(all_of(c(site_col, date_col, response, predictors))) %>% na.omit()

  # 2) Build blocked folds (weekly)
  folds <- make_weekly_folds(df, k = k, date_col = date_col)

  # 3) caret control: use our folds + save OOF preds for by-site metrics
  ctrl <- trainControl(
  method          = "cv",
  number          = k,
  index           = folds$index,
  indexOut        = folds$indexOut,
  verboseIter     = FALSE,
  savePredictions = "final",
  allowParallel   = TRUE      
)

  # 4) Default tuning grids if not provided
  if (is.null(tune_grid)) {
    if (model_type == "rfr") {
      tune_grid <- rf_grid_refined(length(predictors))
    } else {
      tune_grid <- xgb_grid_refined
    }
  }

  # 5) Train
  form <- as.formula(paste(response, "~ ."))
  train_df <- df %>% select(all_of(c(response, predictors)))   # caret formula uses only predictors/response

  if (model_type == "rfr") {
  if (is.null(tune_grid)) tune_grid <- rf_grid_refined(length(predictors))
    fit <- train(
  form,
  data       = train_df,
  method     = "ranger",
  trControl  = ctrl,
  tuneGrid   = tune_grid,
  metric     = "RMSE",
  importance = "permutation",
  num.trees  = 500,
  num.threads = 1            
)
  } else {
    fit <- train(
  form,
  data      = train_df,
  method    = "xgbTree",
  trControl = ctrl,
  tuneGrid  = tune_grid,
  metric    = "RMSE",
  nthread   = 1            
)
  }

  # 6) Out-of-fold predictions joined back to Site for by-site metrics
  # caret's $pred stores CV preds; need the matching rows' Site values
 # --- OOF predictions: align to original rows via caret's rowIndex
if (!"rowIndex" %in% names(fit$pred)) {
  stop("caret did not return 'rowIndex' in fit$pred; cannot align OOF predictions to Site/Date.")
}

oof <- fit$pred %>%
  mutate(
    Site = df[[site_col]][rowIndex],
    Date = df[[date_col]][rowIndex]
  ) %>%
  transmute(Site, Date, obs = obs, pred = pred)

  # 7) Metrics
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # 8) Variable importance
  vimp <- varImp(fit)$importance %>%
    rownames_to_column("Variable") %>%
    arrange(desc(Overall))

  list(
    response     = response,
    model_type   = model_type,
    predictors   = predictors,
    best_params  = fit$bestTune,
    cv_results   = fit$results,
    metrics_overall = mets$overall,
    metrics_by_site = mets$by_site,
    var_importance  = vimp,
    oof_predictions = oof,
    fit_object      = fit
  )
}
```

**Build datasets for each predictor set**
*- Set A (All): exclude Hideaway & Cooper*
*- Set B (Remote-only): include all 7 sites; drop WL/GWT/GWL vars*
```{r}
sites_exclude_all <- c("Hideaway","Cooper")

prep_dataset <- function(df, response_key, mode = c("all","remote")) {
  mode <- match.arg(mode)
  response_col <- resp_cols[[response_key]]

  df2 <- df
  if (mode == "all") {
    df2 <- df2 %>% filter(!(Site %in% sites_exclude_all))
  }
  preds <- get_predictors(df2, response_col, mode = mode)
  list(data = df2, response = response_col, predictors = preds)
}
```

**Run global models for all responses and both predictor sets**
*(XGB and RFR; 5-fold blocked CV)*
```{r}
run_all <- function() {
  results <- list()

  for (resp_key in c("max","mean","range")) {
    # Select the right data frame for this response
    df <- switch(resp_key,
                 max   = max_df,
                 mean  = mean_df,
                 range = range_df)

    # --- Set A: All predictors (5 sites)
    A <- prep_dataset(df, resp_key, mode = "all")
    # XGB
    results[[paste(resp_key,"A_xgb",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"A_rfr",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "rfr", k = 5
    )

    # --- Set B: Remote-only (7 sites)
    B <- prep_dataset(df, resp_key, mode = "remote")
    # XGB
    results[[paste(resp_key,"B_xgb",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"B_rfr",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "rfr", k = 5
    )
  }

  results
}
```

***RUN THE MODELS***
```{r}

cl <- makeCluster(parallel::detectCores() - 1)
registerDoParallel(cl)

results_refined <- run_all()

stopCluster(cl)
```


**Evaluating results**
```{r}
names(results_refined)

# Overall metrics
metrics_overall <- imap_dfr(results_refined, ~ .x$metrics_overall, .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg)

# By-site metrics (with OOF counts)
metrics_by_site <- imap_dfr(results_refined, function(x, nm){
  n_by_site <- x$oof_predictions %>% count(Site, name = "n_oof")
  x$metrics_by_site %>%
    left_join(n_by_site, by = "Site") %>%
    mutate(model_id = nm)
}, .id = NULL) %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg,Site)

# Best params in tidy long form (works for both algs)
best_params_long <- imap_dfr(results_refined, ~{
  tibble::as_tibble(.x$best_params) %>% mutate(model_id = .y)
}, .id = NULL) %>%
  pivot_longer(
    -model_id,
    names_to = "param",
    values_to = "value",
    values_ptypes    = list(value = character()),   # <- force char
    values_transform = list(value = as.character)
  ) %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response, set, alg, param, value)

# “Edge of grid?” check (do we need a second pass?)
edge_check <- imap_dfr(results_refined, function(x, nm){
  bt <- x$best_params
  grid <- x$cv_results
  if (is.null(bt) || is.null(grid) || nrow(grid) == 0) {
    return(tibble(model_id = nm, param = NA_character_, value = NA_character_, at_edge = NA))
  }
  ps <- intersect(names(bt), names(grid))
  purrr::map_dfr(ps, function(p){
    col <- grid[[p]]
    if (is.factor(col)) col <- as.character(col)
    val <- bt[[p]]
    at_edge <- if (is.numeric(col)) {
      suppressWarnings(as.numeric(val)) %in% range(col, na.rm = TRUE)
    } else {
      u <- sort(unique(as.character(col)))
      if (length(u) <= 1) FALSE else as.character(val) %in% c(u[1], u[length(u)])
    }
    tibble(model_id = nm, param = p, value = as.character(val), at_edge = at_edge)
  })
}) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response, set, alg, param, value, at_edge)

# Now check after filtering
edge_hits <- edge_check %>% filter(isTRUE(at_edge))

if (nrow(edge_hits) == 0) {
  message("No parameters hit grid edges.")
} else {
  message("Some parameters hit grid edges.")
}

edge_hits


# Top variables (internal importance)
top_vars <- function(x, k=10) x$var_importance %>% dplyr::slice_max(Overall, n=k)
top_vars(results_refined$max_A_rfr, 10)
top_vars(results_refined$max_A_xgb, 10)

output_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results"
readr::write_csv(metrics_overall, file.path(output_dir, "metrics_overall.csv"))
readr::write_csv(metrics_by_site, file.path(output_dir, "metrics_by_site.csv"))
readr::write_csv(best_params_long,    file.path(output_dir, "best_params_with_cv_metrics.csv"))

```

*k-fold diagnostic*
```{r}
# Fold-level RMSE variability for each model
fold_disp <- purrr::imap_dfr(results_refined, ~{
  res <- .x$fit_object$pred %>%
    group_by(Resample) %>%
    summarise(RMSE = caret::RMSE(pred, obs), .groups = "drop")
  res$model_id <- .y
  res
}) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  group_by(response, set, alg) %>%
  summarise(
    mean_RMSE = mean(RMSE),
    sd_RMSE   = sd(RMSE),
    cv        = sd_RMSE / mean_RMSE,  # coefficient of variation
    .groups = "drop"
  ) %>%
  arrange(desc(cv))

fold_disp
#These look very stable — all CVs are well below 0.15, which means k = 5 is giving consistent fold-level RMSE across all models.
```

*variable importance (for possibly reconsidering including some variables)*
```{r}
get_top_importance <- function(model_list, top_n = NULL) {
  purrr::imap_dfr(model_list, ~{
    vimp <- .x$var_importance %>%
      mutate(model_id = .y)
  })
}

all_vimp <- get_top_importance(results_refined)

# Rank importance within each model
all_vimp <- all_vimp %>%
  group_by(model_id) %>%
  mutate(rank = rank(-Overall, ties.method = "min")) %>%
  ungroup()

# Predictors consistently ranked low
low_ranked <- all_vimp %>%
  group_by(Variable) %>%
  summarise(mean_rank = mean(rank), max_rank = max(rank), .groups = "drop") %>%
  arrange(desc(mean_rank))

low_ranked

# since the models are currently stable, i'm going to just keep all variables for now.

```

**Global lme models**

*setting up VIF function, AR(1) correlation structure*
```{r}
library(nlme)
library(car)        
library(performance)
library(broom)

# Iteratively drop the highest-VIF predictor until all VIF <= thresh
vif_prune <- function(train_df, response, preds, thresh = 5) {
  keep <- preds
  dropped <- character(0)
  # guard: remove constant / near-constant columns
  nzv <- keep[ sapply(keep, function(v) sd(train_df[[v]], na.rm = TRUE) == 0) ]
  if (length(nzv)) {
    keep <- setdiff(keep, nzv)
    dropped <- c(dropped, nzv)
  }
  repeat {
    if (length(keep) == 0L) break
    fm <- as.formula(paste(response, "~", paste(keep, collapse = "+")))
    # try-catch in case of singularities
    fit_lm <- try(lm(fm, data = train_df), silent = TRUE)
    if (inherits(fit_lm, "try-error")) break
    v <- suppressWarnings(car::vif(fit_lm))
    vmax <- max(v, na.rm = TRUE)
    if (is.infinite(vmax) || is.nan(vmax)) break
    if (vmax <= thresh) break
    drop_this <- names(v)[which.max(v)]
    keep <- setdiff(keep, drop_this)
    dropped <- c(dropped, drop_this)
  }
  list(keep = keep, dropped = dropped)
}

# Build AR(1) correlation structure on a Date column
make_cor_ar1 <- function(df, date_col = "Datetime", site_col = "Site") {
  # nlme wants a numeric time; use days since origin
  corAR1(form = as.formula(paste0("~ as.numeric(", date_col, ") | ", site_col)))
}
```

*extra functions for running LME and evaluating VIF*
```{r}
# --- helper to fit lme safely (avoids scoping issue with 'fixed' argument)
fit_lme_safe <- function(fixed_formula, data, site_col, date_col) {
  do.call(nlme::lme, list(
    fixed       = fixed_formula,
    random      = as.formula(paste0("~ 1 | ", site_col)),
    correlation = make_cor_ar1(data, date_col = date_col, site_col = site_col),
    data        = data,
    method      = "REML",
    na.action   = na.omit,
    control     = nlme::lmeControl(msMaxIter = 200, msMaxEval = 400)
  ))
}

# --- safe VIF (long) utility used later to build the VIF log
safe_vif_long <- function(df, response, preds) {
  preds <- unique(preds)
  preds <- preds[!is.na(preds)]
  preds <- preds[preds %in% names(df)]
  if (length(preds) == 0L) {
    return(tibble::tibble(predictor = character(0), vif = numeric(0)))
  }
  fm <- as.formula(paste(response, "~", paste(preds, collapse = "+")))
  v <- try(suppressWarnings(car::vif(stats::lm(fm, data = df))), silent = TRUE)
  if (inherits(v, "try-error")) {
    return(tibble::tibble(predictor = preds, vif = NA_real_))
  }
  if (is.matrix(v)) v <- diag(v) # handle GVIF matrix case
  v <- as.numeric(v)
  v[is.na(v)] <- 1.0             # single-predictor fallback
  tibble::tibble(predictor = preds, vif = v)
}

```


*LME model training fxn*
```{r}
tidy_lme_fixed <- function(fit) {
  tt <- summary(fit)$tTable
  tibble::tibble(
    term      = rownames(tt),
    estimate  = as.numeric(tt[, "Value"]),
    std.error = as.numeric(tt[, "Std.Error"]),
    df        = as.numeric(tt[, "DF"]),
    statistic = as.numeric(tt[, "t-value"]),
    p.value   = as.numeric(tt[, "p-value"])
  )
}


train_lme_global <- function(data,
                             response,
                             predictors,
                             k = 5,
                             site_col = "Site",
                             date_col = "Datetime",
                             vif_thresh = 5) {

  # Keep only needed cols; drop rows with NA in response/preds
  base_df <- data %>%
    dplyr::select(all_of(c(site_col, date_col, response, predictors))) %>%
    tidyr::drop_na()

  # Build weekly folds (your existing helper)
  folds <- make_weekly_folds(base_df, k = k, date_col = date_col)

  # Storage
  oof_list <- vector("list", k)
  fold_kept_preds <- vector("list", k)
  fold_dropped_preds <- vector("list", k)

  for (i in seq_len(k)) {
    val_idx <- folds$indexOut[[i]]
    trn_idx <- folds$index[[i]]

    train_df <- base_df[trn_idx, , drop = FALSE]
    val_df   <- base_df[val_idx, , drop = FALSE]

    # Per-fold VIF pruning (no leakage)
    vp <- vif_prune(train_df, response, predictors, thresh = vif_thresh)
    kept <- vp$keep

          # Build fixed formula
    if (length(kept) > 0) {
      fixed_formula <- reformulate(termlabels = kept, response = response)
    } else {
      fixed_formula <- as.formula(paste(response, "~ 1"))
    }

    # Fit via do.call() so 'fixed' is the actual formula, not a dangling symbol
    fit <- fit_lme_safe(fixed_formula, train_df, site_col, date_col)


    # Predict on validation fold.
    # Use level=1 to include site random intercepts (sites exist in train due to week blocking)
    val_x <- val_df %>% dplyr::select(all_of(c(site_col, date_col, response, kept)))
    val_x <- tidyr::drop_na(val_x)  # ensure no NA for predictors used in this fold

    preds <- predict(fit, newdata = val_x, level = 1)

    oof_list[[i]] <- tibble(
      Site = val_x[[site_col]],
      Date = val_x[[date_col]],
      obs  = val_x[[response]],
      pred = as.numeric(preds)
    )

    fold_kept_preds[[i]]    <- kept
    fold_dropped_preds[[i]] <- vp$dropped
  }

  # Combine OOF; compute metrics
  oof <- bind_rows(oof_list)
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # Fit a final model on full data (with VIF pruning done on full data for reporting)
  vp_full <- vif_prune(base_df, response, predictors, vif_thresh)
  kept_full    <- if (length(vp_full$keep) > 0) vp_full$keep else character(0)
  dropped_full <- vp_full$dropped

  if (length(kept_full) > 0) {
    fixed_formula_full <- reformulate(termlabels = kept_full, response = response)
  } else {
    fixed_formula_full <- as.formula(paste(response, "~ 1"))
  }

  fit_full <- fit_lme_safe(fixed_formula_full, base_df, site_col, date_col)

  # Extract fixed effects table
  coefs <- tidy_lme_fixed(fit_full)

  # AR(1) parameter (phi)
  phi <- tryCatch(
    as.numeric(coef(fit_full$modelStruct$corStruct, unconstrained = FALSE)),
    error = function(e) NA_real_
  )

  list(
    response          = response,
    model_type        = "lme",
    predictors_init   = predictors,
    vif_thresh        = vif_thresh,
    fold_kept         = fold_kept_preds,
    fold_dropped      = fold_dropped_preds,
    kept_full         = kept_full,
    dropped_full      = dropped_full,
    metrics_overall   = mets$overall,
    metrics_by_site   = mets$by_site,
    oof_predictions   = oof,
    fit_object        = fit_full,
    fixed_effects     = coefs,
    ar1_phi           = phi
  )
}
```


*model running fxn for each set*
```{r}
run_all_lme <- function(k = 5, vif_thresh = 5) {
  results <- list()
  for (resp_key in c("max","mean","range")) {
    df <- switch(resp_key, max = max_df, mean = mean_df, range = range_df)

    # --- Set A: All predictors (5 sites)
    A <- prep_dataset(df, resp_key, mode = "all")
    results[[paste(resp_key, "A_lme", sep = "_")]] <- train_lme_global(
      data = A$data, response = A$response, predictors = A$predictors,
      k = k, vif_thresh = vif_thresh
    )

    # --- Set B: Remote-only (7 sites)
    B <- prep_dataset(df, resp_key, mode = "remote")
    results[[paste(resp_key, "B_lme", sep = "_")]] <- train_lme_global(
      data = B$data, response = B$response, predictors = B$predictors,
      k = k, vif_thresh = vif_thresh
    )
  }
  results
}

```

*Run models and export results*
```{r}
# Run
results_lme <- run_all_lme(k = 5, vif_thresh = 5)

# Overall metrics table
metrics_overall_lme <- purrr::imap_dfr(results_lme, ~ .x$metrics_overall, .id = "model_id") %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  dplyr::mutate(alg = "lme") %>%
  dplyr::relocate(response, set, alg)

# By-site metrics (with OOF counts)
metrics_by_site_lme <- purrr::imap_dfr(results_lme, function(x, nm){
  n_by_site <- x$oof_predictions %>% count(Site, name = "n_oof")
  x$metrics_by_site %>% left_join(n_by_site, by = "Site") %>% mutate(model_id = nm)
}, .id = NULL) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  dplyr::mutate(alg = "lme") %>%
  dplyr::relocate(response, set, alg, Site)

# Fixed effects & AR(1)
fixed_effects_lme <- purrr::imap_dfr(results_lme, ~{
  .x$fixed_effects %>% mutate(model_id = .y, ar1_phi = .x$ar1_phi)
}, .id = NULL) %>%
  tidyr::separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  dplyr::mutate(alg = "lme") %>%
  dplyr::relocate(response, set, alg, ar1_phi)

# Save
output_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results"
readr::write_csv(metrics_overall_lme, file.path(output_dir, "lme_metrics_overall.csv"))
readr::write_csv(metrics_by_site_lme, file.path(output_dir, "lme_metrics_by_site.csv"))
readr::write_csv(fixed_effects_lme,  file.path(output_dir, "lme_fixed_effects.csv"))

```

*VIF table*
```{r}
# ---- Build ONE long VIF log file (before/after + dropped) ----
# Uses full-data frames to compute VIFs consistently per (response, set)
vif_log_long <- purrr::imap_dfr(results_lme, function(res, model_id) {
  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  response <- parts[1]
  set_key  <- parts[2]
  alg      <- "lme"

  # Recreate the full dataset used for this response+set (same logic as your runners)
  df_src <- switch(response, max = max_df, mean = mean_df, range = range_df)
  ps <- prep_dataset(df_src, response, mode = ifelse(set_key == "A", "all", "remote"))
  full_df <- ps$data

  # BEFORE: VIF on all candidate predictors (predictors_init)
  vif_before <- safe_vif_long(full_df, res$response, res$predictors_init) %>%
    dplyr::mutate(stage = "before")

  # AFTER: VIF on kept_full (if none kept, produce empty row set)
  vif_after <- safe_vif_long(full_df, res$response, res$kept_full) %>%
    dplyr::mutate(stage = "after")

  # DROPPED: original VIFs for dropped vars (from 'before')
  dropped_info <- vif_before %>%
    dplyr::filter(predictor %in% res$dropped_full) %>%
    dplyr::mutate(stage = "dropped")

  dplyr::bind_rows(vif_before, vif_after, dropped_info) %>%
    dplyr::mutate(
      response = response,
      set      = set_key,
      alg      = alg,
      dropped  = stage == "dropped"
    ) %>%
    dplyr::select(response, set, alg, stage, predictor, vif, dropped)
})

# Save single long file
readr::write_csv(vif_log_long, file.path(output_dir, "lme_vif_log_long.csv"))

```

*Comparing variable importances across all models*
```{r}
# Variable importance for ALL models (RFR, XGB, LME) via permutation (RMSE increase)

library(vip)

options(vip.smaller_is_better = TRUE)

# ---- metric + generic predict wrappers ----
rmse_loss <- function(truth, estimate) sqrt(mean((truth - estimate)^2, na.rm = TRUE))
pred_wrap_caret <- function(object, newdata) predict(object, newdata = newdata)

# For LME we must re-attach grouping/time columns for level=1 predictions
# use fixed-effects only (level = 0) to avoid grouping issues during permutation
make_lme_pred_wrapper <- function(level = 0) {
  function(object, newdata) {
    predict(object, newdata = newdata, level = level)
  }
}

# ---- reuse dataset builder to reconstruct training frames ----
# model_id for ML looks like "max_A_xgb" or "mean_B_rfr"
get_train_frame <- function(model_id) {
  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  response_key <- parts[1]  # "max" | "mean" | "range"
  set_key      <- parts[2]  # "A" | "B"
  df_src <- switch(response_key, max = max_df, mean = mean_df, range = range_df)
  ps <- prep_dataset(df_src, response_key, mode = ifelse(set_key == "A", "all", "remote"))
  list(df = ps$data, response = ps$response, predictors = ps$predictors)
}

set.seed(42)

# Permutation importance for ML models (caret: XGB/RFR) 
perm_vi_ml <- purrr::imap_dfr(results_refined, function(res, model_id) {
  td <- get_train_frame(model_id)
  combo <- td$df %>%
    dplyr::select(dplyr::all_of(c(td$response, td$predictors))) %>%
    tidyr::drop_na()
  train_y <- combo[[td$response]]
  train_x <- combo %>% dplyr::select(dplyr::all_of(td$predictors))

  vi_perm <- vip::vi_permute(
    object            = res$fit_object,   # caret::train
    train             = train_x,
    target            = train_y,
    metric            = rmse_loss,
    pred_wrapper      = pred_wrap_caret,
    nsim              = 20,
    keep              = TRUE,
    parallel          = FALSE,
    smaller_is_better = TRUE
  )

  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  tibble::tibble(
    response   = parts[1],
    set        = parts[2],
    alg        = parts[3],                # "xgb" or "rfr"
    Variable   = vi_perm$Variable,
    Importance = vi_perm$Importance,
    Stderr     = vi_perm$StDev,
    type       = "permutation"
  )
})

# Permutation importance for LME models (fixed-effects only) 
perm_vi_lme <- purrr::imap_dfr(results_lme, function(res, model_id) {
  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  response_key <- parts[1]
  set_key      <- parts[2]

  # Rebuild the same full dataset used for this response+set
  frame <- get_train_frame(paste(response_key, set_key, "x", sep = "_"))

  kept <- res$kept_full
  if (length(kept) == 0L) {
    return(tibble::tibble(
      response = response_key, set = set_key, alg = "lme",
      Variable = character(0), Importance = numeric(0), Stderr = numeric(0), type = "permutation"
    ))
  }

  # Construct aligned training frame; include Site/Datetime just in case
  combo <- frame$df %>%
    dplyr::select(dplyr::all_of(c("Site", "Datetime", frame$response, kept))) %>%
    tidyr::drop_na()

  train_x <- combo %>% dplyr::select(dplyr::all_of(kept))
  train_y <- combo[[frame$response]]

  # Fixed-effects predictions (level = 0) — avoids random-effects grouping errors
  pred_wrap_lme <- make_lme_pred_wrapper(level = 0)

  vi_perm <- vip::vi_permute(
    object            = res$fit_object,   # nlme::lme
    train             = train_x,
    target            = train_y,
    metric            = rmse_loss,
    pred_wrapper      = pred_wrap_lme,
    nsim              = 20,
    keep              = TRUE,
    parallel          = FALSE,
    smaller_is_better = TRUE
  )

  tibble::tibble(
    response   = response_key,
    set        = set_key,
    alg        = "lme",
    Variable   = vi_perm$Variable,
    Importance = vi_perm$Importance,
    Stderr     = vi_perm$StDev,
    type       = "permutation"
  )
})


# include internal importance for ML models (caret::varImp)
internal_vi_all <- purrr::imap_dfr(results_refined, function(res, model_id) {
  parts <- strsplit(model_id, "_", fixed = TRUE)[[1]]
  res$var_importance %>%
    tibble::as_tibble() %>%
    dplyr::transmute(
      response   = parts[1],
      set        = parts[2],
      alg        = parts[3],
      Variable,
      Importance = Overall,
      Stderr     = NA_real_,
      type       = "internal"
    )
})

# Combine & save 
varimp_perm_all <- dplyr::bind_rows(perm_vi_ml, perm_vi_lme)
varimp_combined <- dplyr::bind_rows(internal_vi_all, varimp_perm_all)

readr::write_csv(varimp_perm_all,  file.path(output_dir, "varimp_permutation_ALL_algorithms.csv"))
readr::write_csv(varimp_combined,  file.path(output_dir, "varimp_internal_vs_permutation_ALL_algorithms.csv"))

# Quick peek: top 10 by permutation for each (response,set,alg)
varimp_perm_all %>%
  dplyr::group_by(response, set, alg) %>%
  dplyr::slice_max(Importance, n = 10, with_ties = FALSE) %>%
  dplyr::ungroup()
```

*Variable-importance plots*
```{r}
library(forcats)
library(stringr)

# 1) Filter to Set A (full predictors)
vi_A <- varimp_perm_all %>%
  dplyr::filter(set == "A") %>%
  dplyr::mutate(
    alg = dplyr::recode(alg,
      "lme" = "LME",
      "rfr" = "RFR",
      "xgb" = "XGB",
      .default = alg
    ),
    response = dplyr::recode(response,
      "max" = "Max",
      "mean" = "Mean",
      "range" = "Range",
      .default = response
    ),
    alg = factor(alg, levels = c("LME","RFR","XGB")),
    response = factor(response, levels = c("Max","Mean","Range"))
  )

# 2) Compute ordering: per-response mean importance across algorithms
mean_imp <- vi_A %>%
  group_by(response, Variable) %>%
  summarise(mean_imp = mean(Importance, na.rm = TRUE), .groups = "drop")

# 3) Keep top-k variables per response by that mean importance (tune top_k)
top_k <- 12
top_vars <- mean_imp %>%
  group_by(response) %>%
  slice_max(mean_imp, n = top_k, with_ties = FALSE) %>%
  ungroup()

# 4) Build plotting frame: join mean_imp so we can order by it
plot_df <- vi_A %>%
  inner_join(top_vars, by = c("response","Variable")) %>%  # keeps only top-k per response
  mutate(
    # Create a per-response key so we can order inside facets consistently
    var_key = paste0(response, "__", Variable)
  )

# 5) Optional cosmetic clamp for tiny negatives (noise near zero)
vi_eps <- 0.00  
plot_df <- plot_df %>%
  mutate(Importance = ifelse(abs(Importance) < vi_eps, 0, Importance))

# 6) Plot — order by mean_imp; show same order across alg facets for a response
p_vi_A <- ggplot(
  plot_df,
  aes(x = fct_reorder(var_key, mean_imp), y = Importance)
) +
  geom_col(width = 0.75) +
  geom_errorbar(aes(ymin = pmax(0, Importance - Stderr),
                    ymax = Importance + Stderr),
                width = 0.2, alpha = 0.5) +
  geom_hline(yintercept = 0, linewidth = 0.4, linetype = "dashed") +
  coord_flip() +
  facet_grid(response ~ alg, scales = "free_y", space = "free_y") +
  scale_x_discrete(labels = function(x) sub("^.*__", "", x)) +  
  labs(
    x = "Predictor",
    y = "Increase in RMSE (°C)"
  ) +
  theme_minimal(base_size = 11) +
  theme(
    panel.grid.minor = element_blank(),
    strip.text = element_text(face = "bold"),
    plot.title = element_text(face = "bold"),
    axis.title.y = element_text(margin = margin(r = 10)),
    panel.spacing = unit(0.5, "lines"),         # space between facet panels
    plot.margin   = margin(10, 10, 10, 10)      # margin around whole plot (top, right, bottom, left)
  )


print(p_vi_A)

# Save
ggsave("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/figures/varimp_global.png",
       p_vi_A, width = 8.5, height = 9, units = "in", dpi = 300, bg = "white")

```

*Pred vs Obs plots
```{r}
# helper: label mapping
resp_lab <- c(max = "Max WT", mean = "Mean WT", range = "Range WT")
alg_lab  <- c(lme = "LME", rfr = "RFR", xgb = "XGB")

# 1) Gather OOF predictions from your stored objects:
#    - results_refined: contains *_A_rfr and *_A_xgb
#    - results_lme:     contains *_A_lme
collect_oof_setA <- function(results_refined, results_lme) {
  # From RF/XGB
  df_ml <- purrr::imap_dfr(results_refined, function(x, nm){
    # keep only Set A models
    parts <- strsplit(nm, "_")[[1]]  # c(response, set, alg)
    if (length(parts) != 3) return(NULL)
    if (parts[2] != "A") return(NULL)
    tibble(
      response = parts[1],
      set      = parts[2],
      alg      = parts[3],
      Site     = x$oof_predictions$Site,
      Date     = x$oof_predictions$Date,
      obs      = x$oof_predictions$obs,
      pred     = x$oof_predictions$pred
    )
  })

  # From LME
  df_lme <- purrr::imap_dfr(results_lme, function(x, nm){
    parts <- strsplit(nm, "_")[[1]]
    if (length(parts) != 3) return(NULL)
    if (parts[2] != "A") return(NULL)
    tibble(
      response = parts[1],
      set      = parts[2],
      alg      = "lme",
      Site     = x$oof_predictions$Site,
      Date     = x$oof_predictions$Date,
      obs      = x$oof_predictions$obs,
      pred     = x$oof_predictions$pred
    )
  })

  bind_rows(df_ml, df_lme) %>%
    mutate(
      response = factor(response, levels = c("max","mean","range"), labels = resp_lab),
      alg      = factor(alg,      levels = c("lme","rfr","xgb"),   labels = alg_lab)
    ) %>%
    arrange(response, alg, Site, Date)
}

oof_A <- collect_oof_setA(results_refined, results_lme)


# Distinct site colors + shapes
site_cols   <- c(Anvil="#D55E00", Butte="#0072B2", Crater="#009E73", Pint="#CC79A7", Sam="#E69F00")
site_shapes <- c(Anvil=16, Butte=17, Crater=15, Pint=18, Sam=8)

# Integer breaks with ~5–7 ticks
int_breaks <- function(lims, target = 6L) {
  lo <- floor(lims[1]); hi <- ceiling(lims[2])
  span <- hi - lo
  step <- max(1L, round(span / target))
  seq(lo, hi, by = step)
}

# One response panel (3 rows by model). Lines ABOVE points. 1:1 aspect. Integer ticks.
make_panel <- function(dat, resp_label, show_strips = TRUE) {
  dd <- dat %>% filter(response == resp_label)

  # Per-response limits with small padding
  lims_raw <- range(c(dd$obs, dd$pred), na.rm = TRUE)
  pad      <- diff(lims_raw) * 0.04
  lims     <- c(lims_raw[1] - pad, lims_raw[2] + pad)

  # Integer ticks (same count-ish across responses)
  brk <- int_breaks(lims, target = 6L)

  # Per-facet 1:1 (so it ALWAYS draws, and appears in the legend)
  ones <- dd %>%
    distinct(alg) %>%
    mutate(slope = 1, intercept = 0)

  p <- ggplot(dd, aes(obs, pred)) +
    # points first
    geom_point(aes(color = Site, shape = Site), alpha = 0.8, size = 2.0, stroke = 0.25) +
    # lines ABOVE points
    geom_abline(
      data = ones,
      aes(slope = slope, intercept = intercept, linetype = "1:1"),
      linewidth = 0.8, alpha = 0.95, inherit.aes = FALSE
    ) +
    geom_smooth(aes(linetype = "Fitted"), method = "lm", se = FALSE,
                linewidth = 1.0, alpha = 1.0) +
    facet_grid(rows = vars(alg)) +
    coord_equal(ratio = 1) +
    scale_x_continuous(limits = range(brk), breaks = brk, minor_breaks = NULL,
                       labels = scales::label_number(accuracy = 1)) +
    scale_y_continuous(limits = range(brk), breaks = brk, minor_breaks = NULL,
                       labels = scales::label_number(accuracy = 1)) +
    scale_color_manual(values = site_cols, name = "Site") +
    scale_shape_manual(values = site_shapes, name = "Site") +
    scale_linetype_manual(values = c("1:1" = "dashed", "Fitted" = "solid"), name = "") +
    labs(x = "Observed (°C)", y = "Predicted (°C)") +
    theme_minimal(base_size = 12) +
    theme(
      panel.grid.minor = element_blank(),
      panel.grid.major = element_line(linewidth = 0.35),
      strip.text.y     = if (show_strips) element_text(face = "bold") else element_blank(),
      strip.background = if (show_strips) element_rect(fill = NA, colour = NA) else element_blank(),
      legend.position  = "bottom",
      panel.spacing    = unit(6, "pt"),
      plot.margin      = unit(c(3, 4, 3, 4), "pt")
    )

  p
}

# Build columns: only the RIGHT column shows the facet (model) labels
p_max   <- make_panel(oof_A, "Max WT",  show_strips = FALSE)
p_mean  <- make_panel(oof_A, "Mean WT", show_strips = FALSE)
p_range <- make_panel(oof_A, "Range WT", show_strips = TRUE)   # <- model labels only here

# Headers
hdr <- function(txt) ggplot() + theme_void() +
  annotate("text", x = 0.5, y = 0.5, label = txt, fontface = "bold", size = 5)

col1 <- wrap_plots(hdr("Max WT"),   p_max,   ncol = 1, heights = c(0.06, 1))
col2 <- wrap_plots(hdr("Mean WT"),  p_mean,  ncol = 1, heights = c(0.06, 1))
col3 <- wrap_plots(hdr("Range WT"), p_range, ncol = 1, heights = c(0.06, 1))

# Combine, collect ONE legend, keep columns fairly tight
combined <- (col1 | col2 | col3) +
  plot_layout(guides = "collect", widths = c(1, 1, 1)) &
  theme(
    legend.position = "bottom",
    panel.spacing.x = unit(6, "pt"),
    plot.margin     = unit(c(2, 2, 2, 2), "pt")
  )

# Save
fig_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/figures"
if (!dir.exists(fig_dir)) dir.create(fig_dir, recursive = TRUE, showWarnings = FALSE)

ggsave(file.path(fig_dir, "obs_vs_pred_plot.png"),
       combined, width = 14, height = 8.8, units = "in", dpi = 300, bg = "white")

```

