---
title: "Global wetland ML models"
author: "Michael Krochta"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

**Set up**
```{r}
library(dplyr)
library(lubridate)
library(caret)
library(ranger)     
library(xgboost)    
library(tibble)
library(stringr)
library(tidyr)
library(purrr)
set.seed(42)        # Reproducibility for folds/shuffles
```

**Loading data**
```{r}
max_df  <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/max_global_data.csv",  stringsAsFactors = FALSE)
mean_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/mean_global_data.csv", stringsAsFactors = FALSE)
range_df <- read.csv("C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/data/range_global_data.csv",stringsAsFactors = FALSE)

# Ensure consistent column names and types
# - Datetime to Date
fix_dates <- function(df) {
  if (!inherits(df$Datetime, "Date")) {
    df$Datetime <- as.Date(df$Datetime, tryFormats = c("%Y-%m-%d","%m/%d/%Y","%m/%d/%y"))
  }
  df
}
max_df   <- fix_dates(max_df)
mean_df  <- fix_dates(mean_df)
range_df <- fix_dates(range_df)

# Response column names (adjust if different)
resp_cols <- list(
  max   = "max_WT",
  mean  = "mean_WT",
  range = "range_WT"
)
```

*Define "Remote-only" exclusion list (field-derived vars)*
```{r}
remote_exclude_exact <- c(
  "WL_daily_mean","WL_daily_max","WL_daily_min","WL_daily_range",
  "WL_mean_2day","WL_mean_3day","WL_mean_7day",
  "WL_max_2day","WL_max_3day","WL_max_7day",
  "WL_min_2day","WL_min_3day","WL_min_7day",
  "GWT_mean","GWT_max","GWT_min","GWT_range",
  "GWT_mean_2day","GWT_mean_3day","GWT_mean_7day",
  "GWT_max_2day","GWT_max_3day","GWT_max_7day",
  "GWT_min_2day","GWT_min_3day","GWT_min_7day",
  "GWL_mean","GWL_max","GWL_min","GWL_range",
  "GWL_mean_2day","GWL_mean_3day","GWL_mean_7day",
  "GWL_max_2day","GWL_max_3day","GWL_max_7day",
  "GWL_min_2day","GWL_min_3day","GWL_min_7day", "Water_area_m2_3d"
)
# Also remove any columns that *start with* these prefixes
remote_exclude_prefix <- c("WL_", "GWT_", "GWL_", "Water")

# Build predictor vector for a given df + mode
get_predictors <- function(df, response_col, mode = c("all", "remote")) {
  mode <- match.arg(mode)
  base <- setdiff(names(df), c(response_col, "Site", "Datetime")) # never model on ID/time
  if (mode == "all") return(base)
  # Remote-only: drop exact matches + prefixed columns if present
  drop_prefixed <- unlist(lapply(remote_exclude_prefix, function(p) grep(paste0("^", p), base, value = TRUE)))
  setdiff(base, unique(c(remote_exclude_exact, drop_prefixed)))
}
```


**CV: Build blocked folds by week**
*Keeps whole weeks together (reduces inflated accuracy from adjacent days)*
```{r}
make_weekly_folds <- function(df, k = 5, date_col = "Datetime") {
  df <- df %>% mutate(.week = floor_date(.data[[date_col]], unit = "week", week_start = 1))
  weeks <- unique(df$.week)
  set.seed(42)
  weeks <- sample(weeks)        # shuffle weeks
  folds <- split(weeks, cut(seq_along(weeks), breaks = k, labels = FALSE))

  index <- list(); indexOut <- list()
  for (i in seq_len(k)) {
    val_weeks <- folds[[i]]
    val_idx <- which(df$.week %in% val_weeks)
    trn_idx <- setdiff(seq_len(nrow(df)), val_idx)
    index[[paste0("Fold", i)]]    <- trn_idx
    indexOut[[paste0("Fold", i)]] <- val_idx
  }
  list(index = index, indexOut = indexOut)
}
```

**Metrics**
```{r}
nse <- function(obs, pred) {
  1 - sum((obs - pred)^2, na.rm = TRUE) / sum((obs - mean(obs, na.rm = TRUE))^2, na.rm = TRUE)
}

# Compute overall + by-site metrics from out-of-fold predictions
compute_metrics <- function(pred_df, truth_col = "obs", pred_col = "pred", site_col = "Site") {
  # Overall
  overall <- tibble(
    RMSE = RMSE(pred_df[[pred_col]], pred_df[[truth_col]]),
    MAE  = MAE (pred_df[[pred_col]], pred_df[[truth_col]]),
    R2   = R2  (pred_df[[pred_col]], pred_df[[truth_col]]),
    NSE  = nse (pred_df[[truth_col]], pred_df[[pred_col]])
  )

  # By-site
  by_site <- pred_df %>%
    group_by(.data[[site_col]]) %>%
    summarise(
      RMSE = RMSE(.data[[pred_col]], .data[[truth_col]]),
      MAE  = MAE (.data[[pred_col]], .data[[truth_col]]),
      R2   = R2  (.data[[pred_col]], .data[[truth_col]]),
      NSE  = nse (.data[[truth_col]], .data[[pred_col]]),
      .groups = "drop"
    ) %>%
    rename(Site = .data[[site_col]])

  list(overall = overall, by_site = by_site)
}
```

**Hyperparameter grids (Stage 1: coarse)**
```{r}
rf_grid_coarse <- function(p) {
  expand.grid(
    mtry          = unique(pmax(1, c(floor(sqrt(p)), floor(p/3), floor(p/2)))),
    splitrule     = "variance",
    min.node.size = c(1, 5, 10),
    stringsAsFactors = FALSE   # <-- add this
  )
}

xgb_grid_coarse <- expand.grid(
  nrounds          = c(200, 500),
  max_depth        = c(3, 6, 9),
  eta              = c(0.3, 0.1),
  gamma            = c(0, 0.1),
  colsample_bytree = c(0.8, 1.0),
  min_child_weight = c(1, 3),
  subsample        = c(0.8, 1.0)
)

```

**Core model trainer (XGB or RFR) with blocked CV + saved OOF predictions**
```{r}
train_model <- function(data,
                        response,
                        predictors,
                        model_type = c("xgb", "rfr"),
                        k = 5,
                        site_col = "Site",
                        date_col = "Datetime",
                        tune_grid = NULL) {

  model_type <- match.arg(model_type)

  # 1) Keep only needed columns; drop rows with NA in response/predictors
  df <- data %>% select(all_of(c(site_col, date_col, response, predictors))) %>% na.omit()

  # 2) Build blocked folds (weekly)
  folds <- make_weekly_folds(df, k = k, date_col = date_col)

  # 3) caret control: use our folds + save OOF preds for by-site metrics
  ctrl <- trainControl(
    method         = "cv",
    number         = k,
    index          = folds$index,
    indexOut       = folds$indexOut,
    verboseIter    = FALSE,
    savePredictions= "final"   # <= we’ll compute metrics from these OOF predictions
  )

  # 4) Default tuning grids if not provided
  if (is.null(tune_grid)) {
    if (model_type == "rfr") {
      tune_grid <- rf_grid_coarse(length(predictors))
    } else {
      tune_grid <- xgb_grid_coarse
    }
  }

  # 5) Train
  form <- as.formula(paste(response, "~ ."))
  train_df <- df %>% select(all_of(c(response, predictors)))   # caret formula uses only predictors/response

  if (model_type == "rfr") {
  if (is.null(tune_grid)) tune_grid <- rf_grid_coarse(length(predictors))
    fit <- train(
      form,
      data      = train_df,
      method    = "ranger",
      trControl = ctrl,
      tuneGrid  = tune_grid,
      metric    = "RMSE",
      importance= "permutation", # more reliable than impurity
      num.trees = 500
    )
  } else {
    fit <- train(
      form,
      data      = train_df,
      method    = "xgbTree",
      trControl = ctrl,
      tuneGrid  = tune_grid,
      metric    = "RMSE"
    )
  }

  # 6) Out-of-fold predictions joined back to Site for by-site metrics
  # caret's $pred stores CV preds; need the matching rows' Site values
 # --- OOF predictions: align to original rows via caret's rowIndex
if (!"rowIndex" %in% names(fit$pred)) {
  stop("caret did not return 'rowIndex' in fit$pred; cannot align OOF predictions to Site/Date.")
}

oof <- fit$pred %>%
  mutate(
    Site = df[[site_col]][rowIndex],
    Date = df[[date_col]][rowIndex]
  ) %>%
  transmute(Site, Date, obs = obs, pred = pred)

  # 7) Metrics
  mets <- compute_metrics(oof, truth_col = "obs", pred_col = "pred", site_col = "Site")

  # 8) Variable importance
  vimp <- varImp(fit)$importance %>%
    rownames_to_column("Variable") %>%
    arrange(desc(Overall))

  list(
    response     = response,
    model_type   = model_type,
    predictors   = predictors,
    best_params  = fit$bestTune,
    cv_results   = fit$results,
    metrics_overall = mets$overall,
    metrics_by_site = mets$by_site,
    var_importance  = vimp,
    oof_predictions = oof,
    fit_object      = fit
  )
}
```

**Build datasets for each predictor set**
*- Set A (All): exclude Hideaway & Cooper*
*- Set B (Remote-only): include all 7 sites; drop WL/GWT/GWL vars*
```{r}
sites_exclude_all <- c("Hideaway","Cooper")

prep_dataset <- function(df, response_key, mode = c("all","remote")) {
  mode <- match.arg(mode)
  response_col <- resp_cols[[response_key]]

  df2 <- df
  if (mode == "all") {
    df2 <- df2 %>% filter(!(Site %in% sites_exclude_all))
  }
  preds <- get_predictors(df2, response_col, mode = mode)
  list(data = df2, response = response_col, predictors = preds)
}
```

**Run global models for all responses and both predictor sets**
*(XGB and RFR; 5-fold blocked CV)*
```{r}
run_all <- function() {
  results <- list()

  for (resp_key in c("max","mean","range")) {
    # Select the right data frame for this response
    df <- switch(resp_key,
                 max   = max_df,
                 mean  = mean_df,
                 range = range_df)

    # --- Set A: All predictors (5 sites)
    A <- prep_dataset(df, resp_key, mode = "all")
    # XGB
    results[[paste(resp_key,"A_xgb",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"A_rfr",sep="_")]] <- train_model(
      data = A$data, response = A$response, predictors = A$predictors,
      model_type = "rfr", k = 5
    )

    # --- Set B: Remote-only (7 sites)
    B <- prep_dataset(df, resp_key, mode = "remote")
    # XGB
    results[[paste(resp_key,"B_xgb",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "xgb", k = 5
    )
    # RFR
    results[[paste(resp_key,"B_rfr",sep="_")]] <- train_model(
      data = B$data, response = B$response, predictors = B$predictors,
      model_type = "rfr", k = 5
    )
  }

  results
}
```

***RUN THE MODELS***
```{r}
results <- run_all()
```

**Evaluating results**
```{r}
names(results)

# Global metrics table
metrics_overall <- map_df(results, ~ .x$metrics_overall %>% mutate(.rows = NULL), .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg)

metrics_overall %>% arrange(response,set,alg)

# By-site metrics (+ how many OOF preds per site)
metrics_by_site <- map_df(results, function(x){
  n_by_site <- x$oof_predictions %>% count(Site, name = "n_oof")
  x$metrics_by_site %>%
    left_join(n_by_site, by = "Site") %>%
    mutate(model_id = paste(x$response, ifelse(length(x$predictors)==0,"", "AorB"), x$model_type, sep="_"))
}, .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_", extra = "drop") %>%
  relocate(response,set,alg,Site)

metrics_by_site %>% arrange(response,set,alg,Site)

# Best hyperparameters (joined with global metrics)
best_params <- map_df(results, ~ tibble::as_tibble(.x$best_params), .id="model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  left_join(metrics_overall, by = c("response","set","alg"))

best_params %>% arrange(response,set,alg)

# “Edge of grid?” check (do we need a second pass?)
edge_check <- map_df(results, function(x){
  bt   <- x$best_params
  grid <- x$cv_results %>% dplyr::select(dplyr::any_of(names(bt))) %>%
    mutate(across(where(is.factor), as.character))

  checks <- map2_lgl(names(bt), bt, function(p, val){
    col <- grid[[p]]
    # If column has only one unique value, edge check is irrelevant
    u <- unique(col)
    if (length(u) <= 1) return(FALSE)

    if (is.numeric(col)) {
      v <- suppressWarnings(as.numeric(val))
      v %in% range(col, na.rm = TRUE)
    } else {
      v <- as.character(val)
      u <- sort(unique(as.character(col)))
      v %in% c(u[1], u[length(u)])
    }
  })

  tibble(param = names(bt),
         value = as.character(unlist(bt, use.names = FALSE)),
         at_edge = checks)
}, .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response, set, alg, param, value, at_edge)

edge_check %>% filter(at_edge)

# Top variables (internal importance)
top_vars <- function(x, k=10) x$var_importance %>% dplyr::slice_max(Overall, n=k)
top_vars(results$max_A_rfr, 10)
top_vars(results$max_A_xgb, 10)

# Set the folder path
output_dir <- "C:/Users/mbkro/OneDrive/Documents/wetland-ml-models/results"

# Save tables
readr::write_csv(metrics_overall, file.path(output_dir, "metrics_overall.csv"))
readr::write_csv(metrics_by_site, file.path(output_dir, "metrics_by_site.csv"))
```

*rerunning rfr to account for adding information on best hyperparameters*
```{r}
# Re-run only RF models to save time
rerun_rfr <- function(){
  out <- list()
  for (resp_key in c("max","mean","range")) {
    df <- switch(resp_key, max=max_df, mean=mean_df, range=range_df)

    A <- prep_dataset(df, resp_key, mode = "all")
    out[[paste(resp_key,"A_rfr",sep="_")]] <- train_model(
      data=A$data, response=A$response, predictors=A$predictors,
      model_type="rfr", k=5
    )

    B <- prep_dataset(df, resp_key, mode = "remote")
    out[[paste(resp_key,"B_rfr",sep="_")]] <- train_model(
      data=B$data, response=B$response, predictors=B$predictors,
      model_type="rfr", k=5
    )
  }
  out
}

rfr_results <- rerun_rfr()
# replace the RF entries in your main results list
results[names(rfr_results)] <- rfr_results

#recheck
best_params_long <- map_df(results, ~{
  bp <- tibble::as_tibble(.x$best_params)
  # Make all columns character so pivot has a single type
  bp %>%
    mutate(across(everything(), as.character)) %>%
    pivot_longer(everything(), names_to = "param", values_to = "value")
}, .id = "model_id") %>%
  separate(model_id, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg,param,value)

best_params_long %>% arrange(response,set,alg,param)

readr::write_csv(best_params_long,file.path(output_dir, "best_params_with_cv_metrics.csv"))
```

*Flag edge hits*
```{r}
edge_hits <- map_df(results, function(x){
  bt   <- x$best_params
  grid <- x$cv_results %>% select(any_of(names(bt))) %>%
    mutate(across(where(is.factor), as.character))

  tibble(
    model_id = paste(x$response, x$predictors[1] %||% "", x$model_type, sep="_"),
    param    = names(bt),
    value    = as.character(unlist(bt, use.names = FALSE)),
    at_edge  = map2_lgl(names(bt), bt, function(p, v){
      col <- grid[[p]]
      u <- sort(unique(col))
      if (length(u) <= 1) return(FALSE)                 # constant -> no edge
      if (is.numeric(col)) {
        as.numeric(v) %in% range(col, na.rm = TRUE)
      } else {
        as.character(v) %in% c(u[1], u[length(u)])
      }
    })
  )
}, .id = "run") %>%
  separate(run, into = c("response","set","alg"), sep = "_") %>%
  relocate(response,set,alg,param,value,at_edge)

edge_hits %>% filter(at_edge)

```


